---
title: "Where Brighton's children would go"
subtitle: "A spatial interaction model of secondary school demand in Brighton & Hove, built entirely from published data — with the Longhill question and the 2028 boundary expansion"
author:
  - name: Adam Dennett
    affiliations:
      - name: Bartlett Centre for Advanced Spatial Analysis, University College London
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 150
    fig-cap-location: bottom
    df-print: kable
    embed-resources: true
    link-external-newwindow: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false

library(tidyverse)
library(sf)
library(leaflet)
library(plotly)
library(knitr)
library(kableExtra)
library(scales)
library(ggrepel)
library(here)

source(here::here("public", "R", "00_open_core.R"))

inp <- readRDS(file.path(PUBLIC_OUT, "open_inputs.rds"))
env <- readRDS(file.path(PUBLIC_OUT, "sensitivity_envelope.rds"))
os  <- readRDS(file.path(PUBLIC_OUT, "open_scenarios.rds"))
bright <- readRDS(file.path(PUBLIC_OUT, "brightopia.rds"))
rec    <- readRDS(file.path(PUBLIC_OUT, "reception_cohort.rds"))
adj    <- readRDS(file.path(PUBLIC_OUT, "adjudicator.rds"))
dep    <- readRDS(file.path(PUBLIC_OUT, "deprivation_open.rds"))

LH   <- "Longhill High School"
PCHS <- "Peacehaven Community School"

theme_set(theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey35"),
        panel.grid.minor = element_blank()))

cat_cols <- c("DS_Varndean" = "#4C72B0", "Hove_Blatch" = "#DD8452",
              "Patcham" = "#55A868", "BACA" = "#C44E52",
              "Longhill" = "#8172B3", "PACA" = "#937860",
              "Peacehaven" = "#DA8BC3")

fmt <- function(x, d = 0) formatC(x, format = "f", digits = d, big.mark = ",")

tbl <- function(df, caption = NULL, ...) {
  # Years are labels, not quantities. Without this, big.mark renders 2026
  # as "2,026" directly beneath a date-like header, which reads as an error.
  yr <- grepl("year|round", names(df), ignore.case = TRUE)
  df[yr] <- lapply(df[yr], function(v)
    if (is.numeric(v)) format(v, trim = TRUE, scientific = FALSE) else v)
  kable(df, caption = caption, format.args = list(big.mark = ","), ...) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left", font_size = 13)
}

dem <- function(a, y) round(inp$demand_ts$state_demand[inp$demand_ts$area == a &
                                                         inp$demand_ts$entry_year == y])
cfill <- function(id, y) {
  r <- os$lh_band %>% filter(grepl(paste0("^", id, "\\."), config), entry_year == y)
  sprintf("%.0f%% (%.0f–%.0f)", 100*pmin(r$central,1), 100*pmin(r$lo,1), 100*pmin(r$hi,1))
}
cnat <- function(id, y) round(os$natural$natural[grepl(paste0("^", id, "\\."), os$natural$config) &
                                                   os$natural$entry_year == y])
```

::: {.callout-note}
## Everything here is built from published data

No pupil-level records were used. Every input is published by a public body
or reproducible from published sources using the code in this repository,
and the build enforces that at runtime. Sources are listed in full in
@sec-sources.

The model has parameters that only individual-level admissions data could pin
down precisely. Rather than assume values for them, this analysis sweeps them
across their full plausible range and reports how often each conclusion holds.
Where a finding survives the whole range, no restricted data is needed to state
it — and where one does not, that is reported too.
:::

# What this is

Brighton & Hove has spent two years arguing about secondary school admissions.
The council's proposals, the objections to them, and the consultation evidence
all rested on claims about what would happen if catchments moved, if a school
shrank, or if children were reallocated. None were tested against a model of
how families in the city actually choose schools.

This document builds that model from published data, and uses it to ask three
questions:

1. **What are Brighton families responding to** when they choose a school?
2. **Can Longhill High School be sustained** at its current site and size — and
   does relocating it to the top of Elm Grove change that?
3. **What does the 2028 boundary expansion do**, when Brighton & Hove takes in
   East Saltdean, Telscombe and Peacehaven?

## Who wrote this, and why these methods {#sec-provenance}

I am Professor of Urban Analytics at UCL's Centre for Advanced Spatial Analysis. Spatial interaction models — the class of model this report is built on — have been a central part of my research for around fifteen years, and I have published on their formulation, calibration and application to population flows.

Three pieces are directly relevant to what is done here, and are offered so that the methods can be checked against the literature rather than taken on trust:

- Rowe, F., Lovelace, R. and **Dennett, A.** (2024) [Spatial interaction modelling: a manifesto](https://doi.org/10.4337/9781802203233.00019), in *A Research Agenda for Spatial Analysis*, Edward Elgar. Sets out the modern case for these models, including the reproducibility and open-code standards this report tries to meet.
- **Dennett, A.** and Wilson, A. (2013) [A multilevel spatial interaction modelling framework for estimating interregional migration in Europe](https://doi.org/10.1068/a45398), *Environment and Planning A*, 45(6). The multilevel treatment of nested flows, of which the catchment structure here is a simple case.
- **Dennett, A.** (2018) [Modelling population flows using spatial interaction models](https://discovery.ucl.ac.uk/id/eprint/10052338/), *Australian Population Studies*, 2(2). A practical account of calibrating and validating exactly the production-constrained form used in @sec-method.

This matters for one reason. **The model in this report is not novel, and that is the point.** It is the standard production-constrained spatial interaction model, in continuous use since the 1960s, applied to school admissions in the way it is routinely applied to migration, retail catchments and hospital demand. Its behaviour is well understood, its failure modes are documented, and its calibration is textbook. Nothing here rests on a bespoke method that only its author can evaluate.

What *is* unusual is the constraint: building it entirely from published data, and being explicit at every point about what that costs. Where the published sources cannot support a conclusion, this report says so rather than reaching for one.

::: {.callout-important}
## An offer to the council, and why it matters now

**This analysis is offered as the beginning of a collaboration, not as a criticism.**

Brighton & Hove faces a demographic change that will run for the next fifteen to twenty years. The Year 7 cohort is falling, four wards join the authority in 2028, and the schools affected are not the ones with the most capacity to absorb it. Decisions taken in the next two admissions rounds — one school's admission number here, one catchment boundary there — will compound across that whole period.

The difficulty is that those decisions are currently being taken **one at a time**, each on its own evidence, and often in response to whichever case has been argued most forcefully. A change to Longhill's admission number is considered separately from a change to Dorothy Stringer's, which is considered separately from where a catchment boundary runs, which is considered separately from what happens when Peacehaven joins. Each may be defensible alone. Together they may not add up to a system that works, and nothing in the current process would reveal that.

This document exists to make a different kind of conversation possible: one in which the city's schools are treated as a **single interacting system**, where a change at one school is understood in terms of what it does to every other, across the whole period the demographic change is going to last.

It is built entirely from published data, deliberately, so that anyone can check it. But published data has hard limits, and this report is explicit about every one of them.

**Two kinds of flag appear throughout, and they mean different things:**

- **"This section cannot be done on published data"** — the analysis is absent entirely. Five sections are placeholders of this kind, saying what the records would establish.
- **"This result differs from the calibrated model"** — the analysis *is* here, but the open version and the record-calibrated version disagree. These flags say **which direction the open figure is wrong in, and by roughly how much**, so a reader can discount appropriately rather than take the number at face value.

The second kind matters most. Where this report and the calibrated model diverge, **the open model is almost always the more optimistic of the two about Longhill** — and the flags say so at each point rather than only in a methods appendix.

**What is being asked for is access to the admissions records the council already holds** — preferences, offers, allocation criteria, home postcodes — under the arrangements that already govern research use of pupil data. Nothing new would need collecting. In return, the analysis becomes far sharper: the sections marked below stop being gaps, the largest source of error in this report disappears, and the council gets a tested model of its own system that can be re-run against any proposal before it is consulted on.

The alternative is not that these questions go unanswered. It is that they get answered by whoever argues loudest, using whatever numbers are to hand.
:::

# Method {#sec-method}

## The model

A production-constrained spatial interaction model:

$$T_{ij} = A_i O_i W_j^{\alpha} c_{ij}^{-\beta} \qquad
A_i = \frac{1}{\sum_j W_j^{\alpha} c_{ij}^{-\beta}}$$

where $T_{ij}$ is the flow of children from neighbourhood $i$ to school $j$,
$O_i$ the children living in $i$, $W_j$ school $j$'s attractiveness, $c_{ij}$
the walk/bus journey time, and $A_i$ the balancing factor that places every
child somewhere. Modelled flows are then constrained to school capacity by
iterative proportional fitting, so demand displaced from a full school
cascades to the next-best option rather than vanishing.

**A note on notation.** $c_{ij}$ is the generalised *cost* of travel from $i$
to $j$, measured in minutes of walk and bus journey time routed over the real
network — not a straight-line distance. The symbol is held to that meaning
throughout. The one place a true distance appears is the competing-destinations
term in @sec-brightopia, where $d_{jk}$ is the straight-line separation between
two *schools* in kilometres; that is a different quantity and keeps a different
symbol.

Method after Dennett, *Idiots' Guide to Spatial Interaction Modelling*,
[Part 1](https://rpubs.com/adam_dennett/257231) and
[Part 2](https://rpubs.com/adam_dennett/259068), and the references in
@sec-provenance.

## How $W_j$ was arrived at {#sec-wj}

$W_j$ — how attractive families find each school, net of how far away it is —
is the hardest term in the model to pin down, and the one this bundle can say
least about. It is worth being explicit about that before any result rests on
it.

**The problem.** Attractiveness cannot be read off a school characteristic,
because every candidate is entangled with location. A school with high
Attainment 8 may be sought-after because it teaches well, or because it sits
where affluent families already live. Put a location-correlated variable into a
model that also contains distance and the two fight over the same variation.

**Why this version cannot estimate it.** Resolving that needs individual
preference records — who named which school, from where. With those, $W_j$ can
be estimated freely for each school as a destination fixed effect, letting the
data say how much more often a school is named than its location alone would
predict. Published statistics give only the margins of that table: how many
children live in each area, how many places each school filled. They do not
give the interactions, which is the identification problem set out in
@sec-envelope.

**So it is swept, not chosen.** Four specifications are carried through every
calculation, spanning the plausible range from "nothing distinguishes schools"
to "demand tracks results":

```{r wj-specs}
inp$attract %>%
  transmute(School = name,
            `Equal` = 1,
            `By PAN` = round(W_pan, 2),
            `By first preferences` = round(W_prefs, 2),
            `By Attainment 8` = round(W_att8, 2)) %>%
  arrange(desc(`By first preferences`)) %>%
  tbl(caption = "The four attractiveness specifications. Each is scaled so the city average is 1.")
```

- **Equal** — every school identical. Nobody believes it, but it bounds the
  space and is what @sec-brightopia runs on.
- **Proportional to PAN** — bigger schools attract more, a pure size effect
  with no quality content.
- **Published first preferences** — the council's own factsheets, averaged over
  seven recent rounds. The closest open proxy for revealed demand, and the
  central specification used where a single one is needed.
- **Attainment 8** — demand tracks published results.

::: {.callout-important}
## This is the largest single unknown in the open model

The variance decomposition in @sec-envelope attributes
**`r sprintf("%.0f%%", env$importance$pct[env$importance$term == "w_spec"])`**
of the variation in the answer to which attractiveness specification is used —
more than distance decay
(`r sprintf("%.0f%%", env$importance$pct[env$importance$term == "factor(beta)"])`)
and catchment strength
(`r sprintf("%.0f%%", env$importance$pct[env$importance$term == "factor(gamma)"])`)
combined, several times over.

That is worth stating plainly because it identifies what restricted data would
actually be worth having. Precise travel-time behaviour is **not** the gap —
$\beta$ barely matters. **School-level revealed preference is.** An
origin–destination matrix of Year 7 preferences by area and destination school,
which carries no disclosure risk at catchment level, would replace the widest
band of uncertainty in this document with a measurement.

Two consequences follow for how to read what comes next. Conclusions that hold
across all four specifications are safe to state without restricted data — and
those are reported as such. Conclusions that depend on which specification is
chosen are flagged, and the size of the dependence is given rather than hidden.
:::

Note that first preferences bundle a school's reputation together with where it
happens to be: a school in the middle of the city collects preferences partly
because it is convenient. That is exactly the entanglement the sweep exists to
handle, and it is why the "equal" and "by PAN" specifications are kept in the
range rather than discarded as implausible.

## Where the journey times come from {#sec-network}

Every journey time here is routed by `r5r` over an OpenStreetMap street network and a GTFS bus timetable, rather than measured as straight-line distance. Those two inputs do a great deal of the work, so they are worth showing.

```{r net-map}
#| fig-cap: "The bus network the model routes over. Grey lines are service patterns from the GTFS feed; points are stops. The eastern corridor to Peacehaven, Newhaven and Seaford is what the rebuilt network added."

nm <- readRDS(file.path(PUBLIC_OUT, "network_maps.rds"))

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolylines(data = nm$routes, color = "#555555", weight = 1, opacity = 0.35,
               group = "Bus routes") %>%
  addCircleMarkers(data = nm$stops, radius = 1.5, color = "#2C6E9B",
                   stroke = FALSE, fillOpacity = 0.5,
                   label = ~stop_name, group = "Bus stops") %>%
  addCircleMarkers(data = st_transform(
                     st_as_sf(inp$schools, coords = c("easting", "northing"),
                              crs = 27700), 4326),
                   radius = 5, color = "black", weight = 1,
                   fillColor = "#d95f02", fillOpacity = 0.9,
                   label = ~name, group = "Schools") %>%
  addLayersControl(overlayGroups = c("Bus routes", "Bus stops", "Schools"),
                   options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("Bus stops")
```

The feed carries **`r fmt(nm$n_stops)` stops on `r nm$n_routes` routes**, and the network was built from a merged East and West Sussex OpenStreetMap extract covering `r paste(nm$meta$outcodes, collapse = ", ")` — `r fmt(nm$meta$n_origins)` origin postcodes in all.

### What the network implies for the two sites

```{r net-surface}
#| fig-cap: "Journey time to Longhill from every neighbourhood, at its current site and at the Elm Grove site. Darker is longer."
#| fig-width: 10
#| fig-height: 5

nm$surface %>%
  st_drop_geometry() %>%
  select(zone_e, zone_n, Ovingdean = ovingdean, `Elm Grove` = elm_grove) %>%
  pivot_longer(c(Ovingdean, `Elm Grove`), names_to = "site", values_to = "mins") %>%
  mutate(site = factor(site, c("Ovingdean", "Elm Grove"))) %>%
  ggplot(aes(zone_e, zone_n, colour = mins)) +
  geom_point(size = 2.4) +
  facet_wrap(~ site) +
  scale_colour_viridis_c(option = "magma", direction = -1, name = "Minutes") +
  coord_equal() +
  labs(title = "The same city, two sites",
       subtitle = "Walk-and-bus journey time to Longhill from each neighbourhood",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```

The median neighbourhood is **`r round(median(nm$surface$ovingdean, na.rm = TRUE))` minutes** from the current site and **`r round(median(nm$surface$elm_grove, na.rm = TRUE))` minutes** from Elm Grove. Relocation saves a median of `r round(median(nm$surface$saving, na.rm = TRUE))` minutes, and costs time in `r round(100 * mean(nm$surface$saving < 0, na.rm = TRUE))`% of neighbourhoods — those in the far south-east, which is the trade discussed later.

### Worked journeys

Model outputs are easier to trust when a few of them can be checked against local knowledge. These are real postcodes, routed by the same network that produces every other number here.

```{r net-journeys}
nm$journeys %>%
  transmute(From = place, Postcode = pcd,
            Longhill = round(longhill), `Elm Grove site` = round(elm_grove),
            Stringer = round(ds), Varndean = round(varndean),
            `Cardinal Newman` = round(cardinal_n),
            `Lewes Priory` = round(priory_lewes)) %>%
  tbl(caption = "Walk-and-bus journey time in minutes, morning peak")
```

Three things in that table are worth pausing on, because they explain a great deal of what follows.

**Woodingdean is 17 minutes from Longhill and 68 from Dorothy Stringer** — yet a fifth of the Longhill catchment's children travel out of the city altogether, most of them from here. Distance is plainly not the whole story.

**Whitehawk is 28 minutes from Longhill and 62 from Stringer.** The 2024 boundary change moved most of Whitehawk into Stringer's catchment; the journey time roughly doubles.

**Hove is 60 minutes from Longhill.** This is why no catchment drawn from the centre or west can fill an eastern school: for those families it is not a school, it is an hour each way.

## Why this analysis sweeps rather than calibrates

Published statistics give the *margins* of the origin-destination table — how
many children live in each area, how many places each school fills — but not
the flows between them. Without interaction data, $\beta$ is not identified.

So instead of asserting a value, the analysis runs the model across the whole
plausible range of the three parameters that matter, and reports how often each
conclusion holds:

| Parameter | Swept over | Controls |
| --- | --- | --- |
| $\beta$ | `r min(env$betas)` – `r max(env$betas)` | how sharply demand falls with journey time |
| $W_j$ | `r length(env$w_specs)` specifications | school attractiveness net of location |
| $\gamma$ | `r min(env$gammas)` – `r max(env$gammas)` | strength of catchment priority |

That is `r fmt(nrow(env$envelope))` model runs. Where a single number is needed
to draw a chart, the **central specification** is the midpoint of each range —
$\beta$ = `r os$spec$beta`, $\gamma$ = `r os$spec$gamma`, attractiveness from
published first preferences — chosen for legibility only, with the band from
the sweep reported alongside.

# The system as it stands

```{r map-system}
#| fig-cap: "Brighton & Hove secondary schools and Year 7 demand. Filled areas are the option Z catchments in force from 2026 entry; the dashed outline is the map they replaced. Circles are schools sized by admission number; grey dots are neighbourhoods sized by their modelled 2026 cohort."

catch_sf  <- read_catchments("optionZ") %>% st_transform(4326)
catch_pre <- read_catchments("pre2024") %>% st_transform(4326)

sch_sf <- inp$schools %>%
  left_join(os$choice %>% select(urn, first_pref, prefs_per_place, att8, va), by = "urn") %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700, remove = FALSE) %>%
  st_transform(4326)

zone_sf <- inp$zones %>%
  st_as_sf(coords = c("zone_e", "zone_n"), crs = 27700) %>% st_transform(4326)

pal_c <- colorFactor(unname(cat_cols[catch_sf$catchment]), domain = catch_sf$catchment)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = catch_sf, fillColor = ~pal_c(catchment), fillOpacity = 0.12,
              color = "grey40", weight = 1.5, label = ~catchment,
              group = "Catchments (option Z, from 2026)") %>%
  addPolygons(data = catch_pre, fill = FALSE, color = "black", weight = 1.5,
              dashArray = "5,5", label = ~paste0(catchment, " (pre-2024)"),
              group = "Catchments (pre-2024)") %>%
  addCircleMarkers(data = zone_sf, radius = ~sqrt(Oi) * 1.6, color = "grey30",
                   weight = 0, fillOpacity = 0.35,
                   label = ~paste0(zone, " — ", round(Oi), " children"),
                   group = "Where the children are") %>%
  addCircleMarkers(data = sch_sf, radius = ~sqrt(pan2024) / 1.1, color = "black",
                   weight = 1, fillColor = "#d95f02", fillOpacity = 0.85,
                   label = ~name,
                   popup = ~paste0("<b>", name, "</b><br>PAN 2024: ", pan2024,
                                   "<br>First preferences: ", first_pref,
                                   "<br>Attainment 8: ", round(att8, 1)),
                   group = "Schools") %>%
  addLayersControl(
    overlayGroups = c("Catchments (option Z, from 2026)",
                      "Catchments (pre-2024)",
                      "Where the children are", "Schools"),
    options = layersControlOptions(collapsed = FALSE)) %>%
  hideGroup("Catchments (pre-2024)")
```

The map shows the city's structural problem. The children are concentrated in
the centre and west; two of the eleven schools — Longhill in the far
south-east, and to a lesser extent BACA in the north-east — sit away from where
families live.

::: {.callout-note}
## Which catchment map this uses

The city consulted on secondary catchments in 2024 and determined a new
scheme in March 2025. Under the School Admissions Code an arrangement
determined in February or March of one year first applies to entry the
following September, so the new map — "option Z" — governs entry from 2026.
The filled areas above are option Z; switch on the dashed layer to see the
boundaries it replaced.

The change is confined to east Brighton, and it is not cosmetic. It moves
Kemptown into the Longhill catchment and part of Whitehawk & Marina out of it
into Stringer and Varndean:
`r sum(inp$zones$catch_pre2024 != inp$zones$catch_optionZ)` of this model's
`r nrow(inp$zones)` origin zones change catchment, about
`r sprintf("%.0f%%", 100 * with(inp$pcd_regime, mean(pre2024 != optionZ)))`
of the city's live postcodes.

Because that boundary runs *through* LSOAs, this analysis assigns catchment to
individual postcodes and then splits each LSOA accordingly, rather than
allocating whole LSOAs by their centroid. Assigning wholesale would register
only a fraction of the children the redraw actually moves. Origin zones here
are therefore LSOA × catchment, and each is split on **both** maps at once so
the two regimes sit on identical geography and can be compared directly.
:::

# Brightopia: what the map alone would do {#sec-brightopia}

Before any modelling that requires assumptions, there is a question that
requires none.

Imagine a Brighton in which every secondary school is identical — same size,
same buildings, same staff, same inspection grade, no religious character — and
every child is identical too. The only thing separating the ten schools is
where they stand, and they stand exactly where they stand today. Which school
would each child attend if the only thing that could distinguish them was how
long it takes to get there?

This is the "Brightopia" model from the earlier
[schools_wk3](https://adamdennett.github.io/BH_Schools_2/schools_wk3.html)
work, re-run on the rebuilt routed travel matrix. It is the most robust thing
in this document, because it uses no attainment data, no Ofsted grade, no
preference counts, no catchment rule and no faith criterion. Strip all of that
out and what remains is the geography of the city.

::: {.callout-warning}
## Modelled intake is a market share, not a measure of location

In a production-constrained model the modelled intake of school $j$ is

$$T_j = \sum_i O_i \frac{f(c_{ij})}{\sum_k f(c_{ik})}$$

Reading that in words: $T_j$ is the number of children school $j$ ends up with.
$i$ is a neighbourhood and $O_i$ the children living in it. The top of the
fraction, $f(c_{ij})$, is how attractive the journey from that neighbourhood to
school $j$ is — bigger when the journey is shorter.

**The $k$ on the bottom is the important one.** It is a counter that runs over
*every school in the city, one after another*, adding up the same
journey-attractiveness term for each. So $\sum_k f(c_{ik})$ means "how easy it
is to get from this neighbourhood to schools in general" — not to school $j$,
but to all of them together, $j$ included.

Dividing one by the other turns an absolute quantity into a **share**. The
fraction asks: of all the school-reaching convenience available to this
neighbourhood, what proportion belongs to school $j$? That proportion can fall
in two entirely different ways — because the top gets smaller, meaning school
$j$ is genuinely hard to reach, or because the bottom gets bigger, meaning the
neighbourhood has plenty of other schools within easy reach. An earlier version
of this section reported the result as a single "site index", which treated the
second as though it were the first.

Dorothy Stringer exposes the problem. It sits in the middle of the city with
Varndean 470 metres away and Cardinal Newman and Hove Park close behind. Its
Brightopia intake is low not because children cannot reach it, but because
everyone who can reach it can also reach several alternatives.

The two are separated below: **potential (Hansen) accessibility**, the
distance-weighted count of children within reach, and **Fotheringham's
competing-destinations term**, the accessibility of each school to the other
schools.
:::

```{r brightopia-table}
bright$at_original_beta %>%
  transmute(School = name, PAN = pan2024,
            `Children reachable` = round(access_index),
            `Rivals nearby` = round(compete_index),
            `Brightopia intake` = round(100 * share_index),
            `Mean journey (min)` = round(mean_travel, 1)) %>%
  arrange(`Children reachable`) %>%
  tbl(caption = paste0(
    "Accessibility and competition, separated, at beta = ", bright$beta_original,
    ". Both indices are set to 100 at the city average."))
```

::: {.callout-note}
## How the two columns are calculated

**Children reachable** is potential accessibility, the standard Hansen measure.
For school $j$ it adds up every neighbourhood's children, each discounted by
how long the journey takes:

$$A_j = \sum_i O_i \, c_{ij}^{-\beta}$$

$i$ runs over neighbourhoods, $O_i$ is the children living in one, $c_{ij}$ the
journey time to school $j$, and $\beta$ the distance decay. A child ten minutes
away counts for far more than one an hour away. Nothing about other schools
enters this — it is purely "how many children can get here easily".

**Rivals nearby** is Fotheringham's competing-destinations term. Same shape,
but the sum runs over *schools* rather than neighbourhoods:

$$C_j = \sum_{k \neq j} W_k \, d_{jk}^{-\sigma}$$

Here $k$ counts through the other schools — the $k \neq j$ underneath the sum
means "every school except this one" — $d_{jk}$ is the distance from school $j$
to school $k$, $W_k$ that school's weight, and $\sigma$ how fast the effect
fades with distance. The change of symbol is deliberate: $d_{jk}$ is a
straight-line distance in kilometres between two schools, not a travel-time
cost like $c_{ij}$. Nobody makes the journey from one school to another, so
routing it would be meaningless — what matters is only how clustered the
schools are. In Brightopia every school is identical, so $W_k$ drops
out and the measure becomes pure spatial clustering: with $\sigma = 1$ it
reduces to adding up one-over-the-distance to each of the other nine schools.

Worked through for the two schools that matter here:

| | Nearest neighbour | Distance | Contribution |
|---|---|---|---|
| **Dorothy Stringer** | Varndean | 0.47 km | 1 ÷ 0.47 = **2.14** |
| | Patcham | 1.48 km | 0.67 |
| | Cardinal Newman | 1.67 km | 0.60 |
| | *…and five others* | | *total 5.05* |
| **Longhill** | BACA | 4.71 km | 1 ÷ 4.71 = **0.21** |
| | Varndean | 6.03 km | 0.17 |
| | Dorothy Stringer | 6.31 km | 0.16 |
| | *…and six others* | | *total 1.25* |

Stringer's total is four times Longhill's, and Varndean alone — 470 metres
away — supplies more competition to Stringer than *all nine* of Longhill's
rivals put together. Setting the city average to 100 gives the index values in
the table above.
:::

The two correlate at only
`r sprintf("%.2f", cor(log(bright$at_original_beta$access_index), log(bright$at_original_beta$compete_index)))`,
and the schools at the bottom of the intake column get there by opposite
routes.

**Longhill has the lowest potential accessibility in the city**, at
`r round(bright$at_original_beta$access_index[bright$at_original_beta$name == LH])`,
together with the **lowest** competition term at
`r round(bright$at_original_beta$compete_index[bright$at_original_beta$name == LH])`.
It has almost no rivals nearby, so its low intake cannot be crowding — fewer
children can simply reach it.

**Dorothy Stringer is the opposite**: accessibility
`r round(bright$at_original_beta$access_index[bright$at_original_beta$name == "Dorothy Stringer School"])`,
but the **highest** competition term in the city at
`r round(bright$at_original_beta$compete_index[bright$at_original_beta$name == "Dorothy Stringer School"])`.
Its low modelled intake is crowding, and crowding is exactly what a school
overcomes by being wanted — which Stringer is, filling to its admission number
every year. Longhill cannot escape its position the same way, because no amount
of popularity conjures children into the south-east corner of the city.

Measuring each school against its **own** admission number would mislead in a
further way, because the city has `r fmt(bright$city_places)` places for about
`r fmt(bright$city_children)` children — every school is below its PAN by
construction, and the largest look worst. That is why both indices above are
set to 100 at the city average.

::: {.callout-note}
## A caution about Dorothy Stringer's journey times

Stringer and Varndean are 470 metres apart, yet the routed matrix makes
Stringer slower from most of the city — by a median of about five and a half
minutes. That may be a real difference in bus access, or an artefact of where
`r5r` snapped each school onto the network; it has not been established which.

It is large enough to matter. Giving Stringer exactly Varndean's journey times
raises its Brightopia intake index by roughly 14 points while moving every
other school by no more than three. So a good part of Stringer's apparent
disadvantage is this single discrepancy rather than either location or
competition. Longhill's position is untouched by it, because the anomaly is
between two central schools.
:::

### Would a competing-destinations model fit better?

The separation above is descriptive. Fotheringham's competing-destinations
model (*A new set of spatial-interaction models: the theory of competing
destinations*, Environment and Planning A, 1983) goes further, arguing that the
unadorned gravity model is misspecified when destinations cluster: families
process the choice hierarchically, picking an area and then a school within it,
so omitting a competition term loads spatial structure onto the distance
parameter and biases it.

$$T_{ij} = A_i O_i W_j^{\alpha} C_j^{\delta} c_{ij}^{-\beta}$$

Fitting $\delta$ requires observed origin–destination flows, which is exactly
what published statistics do not provide — the identification problem set out
in @sec-method. So this bundle can construct $C_j$, as it does above, but
cannot estimate its coefficient.

For what it is worth, the same test run on individual preference records finds
$\delta$ statistically indistinguishable from zero, no improvement in fit, and
essentially no change in the distance parameter, across every decay value
tried. The likely reason is that a production-constrained model with origin
fixed effects already handles competition twice over, and that competing-
destinations corrections were built for large interaction systems — inter-city
migration, regional retail catchments — with far more variation in spatial
structure than ten schools inside nine miles. That result cannot be reproduced
from open data, so it is reported here as context rather than relied on: the
descriptive separation above carries the argument on its own.

```{r brightopia-beta}
#| fig-cap: "Longhill in Brightopia, at every distance decay tested. Nothing here depends on how good any school is thought to be."
#| fig-width: 9
#| fig-height: 5

bright$lh_sweep %>%
  select(beta, Ovingdean = fill_now, `Elm Grove` = fill_elm) %>%
  pivot_longer(-beta, names_to = "site", values_to = "fill") %>%
  ggplot(aes(beta, 100 * fill, colour = site)) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1.1) +
  annotate("text", x = max(bright$lh_sweep$beta), y = 104, label = "2024 PAN",
           hjust = 1, size = 3.2, colour = "grey40") +
  scale_colour_manual(values = c(Ovingdean = "#C2413B", `Elm Grove` = "#2C6E9B")) +
  expand_limits(y = 0) +
  labs(title = "Longhill in Brightopia, across every distance decay tested",
       subtitle = "No attainment, no reputation, no catchment, no faith — only the journey",
       x = "Distance decay (beta)", y = "Modelled intake as % of 2024 PAN",
       colour = NULL) +
  theme(legend.position = "bottom")
```

At the decay value the original used, Brightopia gives Longhill
`r round(bright$lh_sweep$modelled_now[bright$lh_sweep$beta == 1.5])` children
where it stands and
`r round(bright$lh_sweep$modelled_elm[bright$lh_sweep$beta == 1.5])` at the top
of Elm Grove. It sits below its 2024 admission number across
`r round(100 * mean(bright$lh_sweep$fill_now < 1))`% of the decay range at
Ovingdean and `r round(100 * mean(bright$lh_sweep$fill_elm < 1))`% at Elm
Grove, and relocation raises its distance-only intake at **every** value of
beta tested. There is no plausible decay parameter at which the current site is
the better placed of the two.

This is worth separating from everything that follows. The rest of this
document depends on assumptions about attractiveness and admissions priority
that open data cannot pin down. This section does not. Whatever else is
uncertain, Longhill is poorly sited relative to where Brighton's children
live, and the Elm Grove site is not.

# Calibration {#sec-calibration-open}

::: {.callout-note}
## This section cannot be done on published data

A spatial interaction model has two parameters that decide everything it says: **how sharply distance deters** a family from choosing a school ($\beta$), and **how much being in catchment shifts a choice** ($\gamma$). Both are estimated by fitting the model to observed flows — how many children actually travel from each neighbourhood to each school.

Published data gives school totals, not flows. There is no open way to estimate either parameter, so this document does not try. Instead it **sweeps** both across their full plausible range and reports which conclusions hold everywhere in that range and which depend on where in it the truth lies (@sec-envelope). That is an honest response to the limitation, and it is why several findings here are stated as ranges rather than numbers.

**With admissions records this becomes a half-day's work.** A Poisson regression of flows on log distance and a catchment indicator, with origin fixed effects, returns both parameters with standard errors. The parallel analysis on records does exactly this and finds a distance decay that is stable across specifications and years — meaning the wide bands in this report are not irreducible uncertainty about Brighton, only uncertainty about what the records would say.

The gain is not academic. Every figure in this report that carries a range would carry a number instead.
:::

# What are families actually choosing? {#sec-choice}

Before modelling anything, a question that can be answered from published data
alone: does demand for a school track its **headline attainment**, or the
**value it adds** to the pupils it teaches?

Both are published. Attainment 8 appears in the DfE performance tables.
Value-added — how a school's results compare with what its intake would
predict — comes from the contextual models in
[school_attainment_tool](https://github.com/adamdennett/school_attainment_tool).
Demand comes from the council's own allocation factsheets.

```{r choice-table}
os$choice %>%
  transmute(School = school, PAN = pan, `First preferences` = first_pref,
            `Prefs per place` = round(prefs_per_place, 2),
            `Attainment 8` = round(att8, 1),
            `Value-added` = round(va, 2)) %>%
  arrange(desc(`Prefs per place`)) %>%
  tbl(caption = "Published demand and published performance")
```

```{r choice-plot}
#| fig-cap: "First preferences per place against the two published measures of school quality. Hover for detail."
#| fig-height: 5

cl <- os$choice %>%
  select(school, prefs_per_place, pan, att8, va) %>%
  pivot_longer(c(att8, va), names_to = "measure", values_to = "value") %>%
  mutate(measure = factor(measure, c("att8", "va"),
                          c("Attainment 8 — the league-table score",
                            "Value-added — the school's own contribution")))

gg <- ggplot(cl, aes(value, prefs_per_place, size = pan,
                     text = paste0(school, "<br>Prefs/place: ",
                                   round(prefs_per_place, 2),
                                   "<br>Value: ", round(value, 2)))) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey65") +
  geom_smooth(aes(group = 1), method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey40", linewidth = 0.5, linetype = "dotted") +
  geom_point(alpha = 0.85, colour = "#1f77b4") +
  facet_wrap(~ measure, scales = "free_x") +
  labs(x = NULL, y = "First preferences per place") +
  theme(legend.position = "none")

ggplotly(gg, tooltip = "text")
```

```{r choice-models}
os$choice_tbl %>%
  transmute(Specification = spec, `R²` = round(r2, 3), Coefficients = terms) %>%
  tbl(caption = "Explaining demand. n = 10, so this is descriptive — no p-values are reported.")
```

::: {.callout-important}
## The finding

Raw Attainment 8 explains
**`r sprintf("%.0f%%", 100*os$choice_tbl$r2[os$choice_tbl$spec == "Attainment 8"])`**
of the variation in how much Brighton families want a school. Value-added — the
measure of what the school actually contributes once its intake is accounted
for — explains
**`r sprintf("%.0f%%", 100*os$choice_tbl$r2[os$choice_tbl$spec == "Value-added"])`**.

Put both in the same model and Attainment 8 stays strongly positive while
value-added turns *negative*.

Families are choosing the school with the higher headline score, not the school
that does more for its pupils. Since headline scores are largely a function of
who a school teaches, the "golden ticket" of a place at a high-scoring school
is mostly a ticket to sit alongside higher-attaining classmates — not to be
better taught.
:::

This is not a criticism of parents. Attainment 8 is published, prominent and
easy to compare; contextual value-added is not published in any form a parent
would encounter. But it does undercut the premise that moving children between
schools transfers a quality benefit. The distributional side of that
question — which children live where — is in @sec-deprivation. Brighton Aldridge Community Academy has
one of the better value-added records in the city and one of the lowest demand
ratios; Patcham High has the weakest value-added and sits close to the city
average for demand.

::: {.callout-note}
## Which lever this leaves unpulled

The companion report *How to Pull the Right Lever* finds that **absence** is by
some distance the most powerful predictor of attainment available to a school
in England — with a coefficient around 2.6 times that of concentrations of
disadvantage — and that Brighton & Hove had the second worst absence rate in
the country in 2024–25.

Set that beside the finding above and the shape of the problem becomes clear.
The measure families sort on, raw Attainment 8, is largely a function of who a
school already teaches. The measure that reflects what a school does with its
intake, value-added, is close to invisible in their choices. And the factor with
the most mechanical advantage over attainment, attendance, plays no part in the
choice process at all — it is not published in any comparable form and does not
enter the decision.

Two consequences follow for the question this document asks. First, a school
that genuinely improves should not expect demand to respond quickly, because
the signal families act on barely registers improvement — which is a hard
constraint on any Longhill recovery plan built around families noticing.
Second, a policy of redistributing children between schools is working on the
lever with less mechanical advantage, and does so by lengthening journeys —
which the same literature associates with *higher* absence, the lever that
matters most. That is the argument of the companion report, and this model's
demand estimates are consistent with it.
:::

# Does the model work?

```{r validation}
#| fig-cap: "Modelled intake against the council's published Year 7 offers, at the central specification."

gg <- ggplot(os$val, aes(observed, modelled,
                         text = paste0(name, "<br>Published: ", round(observed),
                                       "<br>Modelled: ", round(modelled)))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey55") +
  geom_point(size = 3, colour = "#1f77b4", alpha = 0.85) +
  labs(title = sprintf("R² = %.3f, RMSE = %.0f pupils",
                       CalcRSquared(os$val$observed, os$val$modelled),
                       CalcRMSE(os$val$observed, os$val$modelled)),
       x = "Published offers", y = "Modelled intake")

ggplotly(gg, tooltip = "text")
```

```{r validation-table}
os$val %>%
  transmute(School = name, Published = round(observed), Modelled = round(modelled),
            Difference = round(modelled - observed)) %>%
  arrange(desc(abs(Difference))) %>%
  tbl(caption = "Where the open model is wrong, and by how much")
```

The fit is moderate
(R² = `r sprintf("%.2f", CalcRSquared(os$val$observed, os$val$modelled))`), and
one error dominates it: **the model over-predicts Longhill by
`r round(os$val$modelled[os$val$name == LH] - os$val$observed[os$val$name == LH])`
children**, against `r round(os$val$observed[os$val$name == LH])` actually
offered.

That is worth dwelling on, because it is the central limitation of an
open-data model of school choice. The model knows where Longhill is, how big it
is, and how long it takes to reach — and on those facts it should recruit well.
What it cannot see is that Brighton families do not want it. Only
individual-level preference data measures that, and it is not published.

**Every figure below is therefore generous to Longhill.** Where the model says
the school struggles to fill, the real position is worse; where it says
relocation helps, the true gain is smaller.

# Reading the admissions rules instead of inferring them {#sec-rules-open}

::: {.callout-note}
## This section cannot be done on published data

Every model in this document *infers* how the admissions rules work from where children end up. The records contain the answer directly: each offer carries the criterion under which it was made — sibling, catchment, free school meals, distance, looked-after — so the rules can be read rather than reconstructed.

That distinction matters more than it sounds, because inference and reading can disagree. The parallel analysis on records finds one result that no amount of open modelling would have produced: **catchment priority is worth essentially nothing at Longhill.** Almost everyone who names the school is offered a place, whether they live in its catchment or not, because it is undersubscribed. Priority only bites where places are scarce.

If that is right, it changes the policy conclusion entirely. **Redrawing Longhill's catchment cannot work by letting more children in — nobody is being turned away.** It can only work by changing which school families *name*. That is a much weaker and slower mechanism, and any proposal resting on a redraw needs to be honest about which of the two it is relying on.

This report cannot confirm or refute that finding. It is the clearest single example of a conclusion that is invisible in published data, cheap to establish with records, and directly consequential for a decision the city is about to take.
:::

# Do all families choose the same way? {#sec-heterogeneity-open}

::: {.callout-note}
## This section cannot be done on published data

Everything above treats families as responding identically to distance and attractiveness. They do not. Splitting applicants by the deprivation of their home neighbourhood and fitting the model separately to each group tests whether the weight placed on a school's reputation, and the willingness to travel for it, vary across the social gradient.

The parallel analysis finds they vary a great deal — the weight on attractiveness roughly **doubles** from the most to the least deprived third of neighbourhoods, and the difference is strongly significant. Allowing for it lifts the model's fit substantially and reproduces a sorting pattern that a single-parameter model cannot: less deprived families in a shared catchment securing the school they prefer, more deprived families not.

That is the mechanism behind the Hove Park and Blatchington Mill contrast in @sec-deprivation, and it cannot be seen in aggregate data at all — every child in an LSOA looks identical to a model built on published sources. **Any policy aimed at social mix is aimed at this mechanism**, so not being able to measure it is a serious limitation on what this report can say about equity.

Only individual records can separate it. Postcodes alone are not enough; the preference ordering is what carries the signal.
:::

# How good would Longhill have to be? {#sec-required-open}

::: {.callout-note}
## This section cannot be done on published data

A useful way to state the problem in reverse: rather than asking what Longhill's intake would be under some scenario, ask **how much more attractive the school would need to become to fill at a given admission number**, holding its location and everything else constant. That converts an abstract question about reputation into a number on the same scale as the attainment measures the model already uses.

It needs a calibrated model, which needs records (@sec-calibration-open). With one, the answer is a single figure per admission number, and it can be compared against how far schools have actually moved on that scale in practice — which tells you whether the required improvement is ambitious, or simply outside anything ever observed.

That comparison is the honest test of whether "improve the school" is a viable alternative to changing its size or location. This report cannot make it.
:::

# A shrinking cohort

```{r demand-2035}
#| fig-cap: "Projected state-sector Year 7 demand. Solid points are cohort-aged from children already counted in the ONS mid-2022 estimates; open points from 2034 are extrapolated."

inp$demand_ts %>%
  filter(entry_year >= 2026) %>%
  ggplot(aes(entry_year, state_demand, colour = area)) +
  geom_line(linewidth = 0.9) +
  geom_point(aes(shape = extrapolated), size = 2.6, fill = "white") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 21), guide = "none") +
  scale_colour_manual(values = c("Brighton & Hove" = "#4C72B0",
                                 "Expansion area" = "#DD8452")) +
  scale_x_continuous(breaks = seq(2026, 2035, 1)) +
  expand_limits(y = 0) +
  labs(title = "The cohort keeps shrinking, expansion or not",
       subtitle = "ONS small area estimates, aged forward to Year 7 entry",
       x = "Year 7 entry year", y = "Children", colour = NULL) +
  theme(legend.position = "bottom")
```

A second, independently grounded projection from Reception offers is set out in @sec-reception.

A child aged $a$ in the ONS mid-2022 estimates enters Year 7 in
$2022 + (11 - a)$, so entry years to 2033 come from children already born and
counted. Only 2034 and 2035 are extrapolated.

Brighton & Hove's Year 7 cohort falls from about `r fmt(dem("Brighton & Hove", 2026))`
in 2026 to `r fmt(dem("Brighton & Hove", 2035))` in 2035 — a drop of
**`r round(100*(1 - dem("Brighton & Hove", 2035)/dem("Brighton & Hove", 2026)))`%**.
The expansion area adds `r dem("Expansion area", 2026)` children in 2026 and
`r dem("Expansion area", 2035)` by 2035.

## A second projection, from children already in school {#sec-reception}

The projection above counts children *in the population*. What schools need is
children who will take a *state-sector* place, and the gap between the two —
independent schools, home education, migration — is large in Brighton.

There is a better-grounded alternative in the published record, and it is the
method already used in
[brighton_and_hove_allocations](https://adamdennett.github.io/BH_Schools_2/brighton_and_hove_allocations.html):
compare Reception offers with the Year 7 offers made to the **same cohort**
seven years later. A child offered a Reception place in 2014 was offered a Year
7 place in 2021. The ratio between them is a measured survival rate, not an
assumption.

```{r rec-cohort}
rec$cohort %>%
  transmute(`Reception year` = cohort_year,
            `Reception offers` = rec_offers,
            `Year 7 year` = y7_year,
            `Year 7 offers` = y7_offers,
            Change = change,
            `Survival` = sprintf("%.3f", survival)) %>%
  tbl(caption = "The same cohort, at ages 4 and 11")
```

The rate is remarkably steady: `r sprintf("%.3f", rec$survival_recent)` across
the five most recent cohorts, with a standard deviation of
`r sprintf("%.3f", rec$survival_sd)` and no significant trend (p =
`r sprintf("%.2f", rec$survival_trend_p)`). About
`r round(100 * (1 - rec$survival_recent))`% of a Reception cohort has left the
city's state sector by Year 7.

Applying it to cohorts already sitting in primary school projects Year 7 demand
to **2033 entry without extrapolating a single birth**:

```{r rec-proj}
rec$comparison %>%
  transmute(`Entry year` = entry_year,
            `From Reception offers` = round(reception_based),
            `From ONS estimates` = round(ons_demand),
            Difference = round(diff),
            `%` = sprintf("%+.1f%%", pct),
            Note = if_else(extrapolated, "ONS figure extrapolated", "")) %>%
  tbl(caption = "Two independent projections of Year 7 demand")
```

The two agree within about 3% for most years, which is reassuring for both. They
diverge most in **2033**, where the Reception-based figure is
`r abs(round(rec$comparison$diff[rec$comparison$entry_year == 2033]))` children
lower — and that is precisely the year where the ONS series is extrapolated
rather than counted. The Reception figure should be preferred there.

## What the council itself projects

```{r rec-council}
rec$council %>%
  select(CatchmentGroup, PAN, entry_year, council) %>%
  pivot_wider(names_from = entry_year, values_from = council) %>%
  rename(Catchment = CatchmentGroup) %>%
  tbl(caption = "Brighton & Hove's own Oct-2025 Year 7 forecast, by catchment")
```

Two things in the council's own forecast deserve more attention than they have
had.

**It plans around a Longhill PAN of `r rec$council_lh_pan`.** The adjudicator
has since confirmed that number as binding for 2026/27, and it is the one used
throughout this document. The 240 that appears alongside it in places is the
figure Longhill carried until 2026, kept for comparison rather than as a live
option.

**Its 2032 figure for the Longhill catchment is
`r rec$council$council[rec$council$CatchmentGroup == "Longhill" & rec$council$entry_year == 2032]`
children** — against that PAN of `r rec$council_lh_pan`, a shortfall of
`r rec$council_lh_pan - rec$council$council[rec$council$CatchmentGroup == "Longhill" & rec$council$entry_year == 2032]`.
Across 2026 to 2032 the council's Longhill line falls
`r round(100 * (rec$council$council[rec$council$CatchmentGroup == "Longhill" & rec$council$entry_year == 2032] / rec$council$council[rec$council$CatchmentGroup == "Longhill" & rec$council$entry_year == 2026] - 1))`%.

::: {.callout-warning}
## Most of that decline is an assumption, not a demographic

The council's forecast reduces each catchment's roll by a catchment-specific
rate for children expected to leave the state-maintained sector. Published in
the same appendix, those rates are: PACA 2.46%, Hove Park / Blatchington Mill
5.13%, Varndean / Dorothy Stringer 4.58%, BACA 6.52%, Patcham 3.03% — and
**Longhill 23.89%**, more than five times the average of the other five. That assumption is tested against published evidence in @sec-outflow.

The Reception cohorts actually feeding the Longhill catchment fall by about
`r if (!is.null(rec$by_catchment)) { lh <- rec$by_catchment[rec$by_catchment$catchment == "Longhill", ]; paste0(abs(round(100 * (lh$rec_offers[lh$entry_year == 2032] / lh$rec_offers[lh$entry_year == 2026] - 1))), "%") } else "20%" `
over the same span. The council's line falls roughly twice as fast, and the
difference is that single leakage rate.

The assumption may well be right — Longhill does lose a great many children to
other schools, which is the entire subject of this document. But it is close to
self-fulfilling as a planning device: a high leakage rate justifies shrinking
the school, and a shrinking school raises leakage. It deserves to be stated as
a judgement and tested, rather than carried through as a parameter.
:::

### Every catchment, side by side {#sec-catchment-projections}

The comparison above is worth doing for all six catchments rather than Longhill alone, because it shows how far Longhill's treatment stands apart. This replicates the cohort-ageing projection from [*Brighton and Hove allocations*](https://adamdennett.github.io/BH_Schools_2/brighton_and_hove_allocations.html#our-own-projection-cohort-aging-from-primary-allocations), which ages each catchment's Reception offers forward seven years to the Year 7 cohort they become.

```{r catch-proj}
# Our internal catchment keys and the council's published labels name the
# same places. They must be harmonised before any join, or the two paired
# catchments silently drop out of every comparison below.
CATCH_LABEL <- c(PACA = "PACA",
                 Hove_Blatch = "Hove Park / Blatchington Mill",
                 Patcham = "Patcham",
                 DS_Varndean = "Varndean / Dorothy Stringer",
                 BACA = "BACA",
                 Longhill = "Longhill",
                 Peacehaven = "Peacehaven",
                 `Religious schools` = "Religious schools")
relabel <- function(x) dplyr::coalesce(unname(CATCH_LABEL[x]), x)

ours_lab <- rec$by_catchment %>% mutate(catchment = relabel(catchment))
council_lab <- rec$council %>%
  transmute(catchment = relabel(CatchmentGroup), entry_year, council)

cp <- ours_lab %>%
  filter(entry_year %in% c(2026, 2029, 2032)) %>%
  select(catchment, entry_year, ours = projected) %>%
  left_join(council_lab, by = c("catchment", "entry_year")) %>%
  filter(!is.na(council))

stopifnot(dplyr::n_distinct(cp$catchment) == 6)

cp %>%
  mutate(txt = sprintf("%d / %d", round(ours), round(council))) %>%
  select(catchment, entry_year, txt) %>%
  pivot_wider(names_from = entry_year, values_from = txt) %>%
  rename(Catchment = catchment) %>%
  tbl(caption = "Children aged forward from Reception offers / the council's own forecast, by catchment and entry year")
```

```{r catch-proj-plot}
#| fig-cap: "Each catchment's Reception cohort carried forward at that catchment's own measured Reception-to-Year-7 ratio (solid) against the council's October 2025 forecast (dashed). Longhill is the catchment where the council's line sits furthest above the cohorts behind it."
#| fig-height: 6

LV <- c("Year 7 offers actually made", "Projected from Reception cohorts",
        "Council forecast, Oct 2025")

pj <- bind_rows(
  rec$observed_by_catchment %>%
    transmute(catchment, entry_year, n = y7_offers, src = LV[1]),
  rec$by_catchment %>%
    filter(entry_year > max(rec$observed_by_catchment$entry_year)) %>%
    transmute(catchment, entry_year, n = projected, src = LV[2]),
  rec$council %>%
    transmute(catchment = CatchmentGroup, entry_year, n = council, src = LV[3])) %>%
  filter(!is.na(n)) %>%
  mutate(catchment = relabel(catchment), src = factor(src, levels = LV))

# Join the observed series to the projection so the line is continuous
# rather than breaking at the handover year.
bridge <- rec$observed_by_catchment %>%
  filter(entry_year == max(entry_year)) %>%
  transmute(catchment = relabel(catchment), entry_year, n = y7_offers,
            src = factor(LV[2], levels = LV))

# Free y-scales otherwise crop the taller panels awkwardly. These are
# headroom values, not data.
ylim_df <- tibble(
  catchment = c("Hove Park / Blatchington Mill", "Varndean / Dorothy Stringer",
                "Religious schools"),
  n = c(625, 700, 600), entry_year = 2026,
  src = factor(LV[1], levels = LV))

bind_rows(pj, bridge) %>%
  ggplot(aes(entry_year, n, colour = src, linetype = src)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.5) +
  geom_blank(data = ylim_df) +
  facet_wrap(~ catchment, ncol = 3, scales = "free_y") +
  scale_colour_manual(values = setNames(c("grey30", "#1b9e77", "#d95f02"), LV)) +
  scale_linetype_manual(values = setNames(c("solid", "solid", "dashed"), LV)) +
  expand_limits(y = 0) +
  labs(title = "What actually happened, what the cohorts imply, and what the council forecasts",
       subtitle = "Year 7 offers by catchment. Projection applies each catchment's own measured Reception-to-Year-7 ratio",
       x = "Year 7 entry year", y = "Children", colour = NULL, linetype = NULL) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2))
```

The pattern is easier to see as the gap itself — the council's forecast expressed as a percentage of the children the Reception cohorts imply.

```{r catch-proj-gap}
#| fig-cap: "The council's forecast as a share of the cohort implied by each catchment's own conversion rate. Above 100% means the council is forecasting more children than the primary cohorts support."

gap <- ours_lab %>%
  select(catchment, entry_year, ours = projected) %>%
  inner_join(council_lab, by = c("catchment", "entry_year")) %>%
  mutate(ratio = 100 * council / ours,
         lh = catchment == "Longhill")

gap %>%
  ggplot(aes(entry_year, ratio, group = catchment, colour = lh)) +
  geom_hline(yintercept = 100, linetype = "dotted", colour = "grey50") +
  geom_line(aes(linewidth = lh)) + geom_point(size = 1.8) +
  scale_colour_manual(values = c(`TRUE` = "#d95f02", `FALSE` = "grey65"),
                      labels = c(`TRUE` = "Longhill", `FALSE` = "Other catchments")) +
  scale_linewidth_manual(values = c(`TRUE` = 1.2, `FALSE` = 0.6), guide = "none") +
  ggrepel::geom_text_repel(
    data = gap %>% filter(entry_year == max(entry_year)),
    aes(label = catchment), size = 3, hjust = 0, direction = "y",
    nudge_x = 0.3, segment.colour = "grey80", show.legend = FALSE) +
  scale_x_continuous(limits = c(2026, 2034)) +
  expand_limits(y = 0) +
  labs(title = "Where the council forecasts more than the primary cohorts support",
       subtitle = "Council forecast as a percentage of the cohort-implied figure",
       x = "Year 7 entry year", y = "Council forecast as % of cohort", colour = NULL) +
  theme(legend.position = "bottom")
```

```{r catch-proj-fall}
#| include: false
fall <- cp %>% group_by(catchment) %>%
  summarise(ours_pct = 100 * (ours[entry_year == 2032] / ours[entry_year == 2026] - 1),
            cou_pct  = 100 * (council[entry_year == 2032] / council[entry_year == 2026] - 1),
            .groups = "drop")
lh_f <- fall %>% filter(catchment == "Longhill")
```

::: {.callout-important}
## The council's forecast is above what Longhill's own primary cohorts support, in every year

The ratio matters more than the raw cohort here, and it is measured rather than assumed. For each catchment it is the Year 7 offers made *by that catchment's own secondaries*, over the Reception offers made in that catchment seven years earlier — so it absorbs leakage, cross-catchment movement and capacity limits in one observed number.

```{r catch-ratio}
rec$by_catchment %>% distinct(catchment, surv_catch) %>%
  transmute(Catchment = catchment, `Reception-to-Year-7 ratio` = round(surv_catch, 3)) %>%
  arrange(`Reception-to-Year-7 ratio`) %>%
  tbl(caption = "Share of each catchment's Reception cohort that reaches that catchment's own secondary schools")
```

**Longhill retains `r sprintf("%.0f", 100 * unique(rec$by_catchment$surv_catch[rec$by_catchment$catchment == "Longhill"]))`% of its Reception children — the lowest in the city, against a city-wide figure of `r sprintf("%.0f", 100 * rec$survival_recent)`%.** Fewer than half the children starting primary school in the Longhill catchment take a Year 7 place at Longhill. Brighton Aldridge, at the other extreme, recruits more than its own catchment produces.

Apply each catchment's own ratio and the council's forecast for Longhill sits **above** what the primary cohorts support, in every projected year: `r round(cp$council[cp$catchment == "Longhill" & cp$entry_year == 2026])` against `r round(cp$ours[cp$catchment == "Longhill" & cp$entry_year == 2026])` in 2026, and `r round(cp$council[cp$catchment == "Longhill" & cp$entry_year == 2032])` against `r round(cp$ours[cp$catchment == "Longhill" & cp$entry_year == 2032])` in 2032. No other catchment shows a gap of that kind.

**This is where the sleepwalking risk actually sits.** The council's forecast already looks pessimistic about Longhill relative to other catchments, and it is *still* not pessimistic enough. Its 23.89% leakage deduction is applied to a cohort figure that assumes Longhill converts its catchment at something like the city-wide rate. It does not, and has not for at least six years.

The calibrated behavioural model reaches the same place independently and goes further: around 79 children by 2035, below both lines above.

Two consequences follow. First, decisions about Longhill's admission number are being taken against a demand figure that the school's own primary catchment does not support. Second, using leakage as the sole adjustment is self-reinforcing as a planning device — a high assumed leakage justifies shrinking the school, and a shrinking school raises leakage — while never asking the question that matters, which is *what would change the conversion rate*.
:::

## The assumption is testable, and the adjudicator published the test {#sec-outflow}

When the Office of the Schools Adjudicator determined objections to the 2026/27
arrangements on 20 October 2025 (`r adj$case`), it required the council to
produce evidence, and published it. One of those tables answers the question
directly: how many children in each catchment are offered a place **outside**
Brighton & Hove.

```{r adj-outside}
adj$outside %>%
  transmute(Catchment = catchment, round,
            v = sprintf("%d (%.0f%%)", children, share)) %>%
  pivot_wider(names_from = round, values_from = v) %>%
  tbl(caption = paste0(
    "Children offered a place outside Brighton & Hove on National Offer Day, ",
    "by home catchment. Percentages are of the city-wide total that year. ",
    "Source: adjudicator's determination, ", format(adj$determined, "%d %B %Y"),
    ", Table 11."))
```

```{r adj-nums}
#| include: false
lh_out <- adj$outside %>% filter(catchment == "Longhill")
lh_25 <- lh_out$children[lh_out$round == "2025/26"]
tot_25 <- lh_out$total[lh_out$round == "2025/26"]
```

**The Longhill catchment accounts for `r round(min(lh_out$share))`–`r round(max(lh_out$share))`% of every child who leaves Brighton & Hove for a school elsewhere.** In the most recent round it was `r lh_25` children of `r tot_25` city-wide. Every other catchment loses single figures.

This is the strongest published evidence in this document, and it changes what
can be said. The council's `r "23.89"`% leakage assumption is not an arbitrary
planning parameter — it is measuring something real and specific, concentrated
almost entirely in one catchment. Set against a Longhill catchment cohort of
roughly 290 children, `r lh_25` leavers is about `r round(100 * lh_25 / 291)`%,
which is the same order as the rate the council applies.

The council's own forecast makes the same point from the other direction:

```{r adj-retention}
adj$retention %>%
  transmute(Catchment = catchment, round, v = sprintf("%.0f%%", retained)) %>%
  pivot_wider(names_from = round, values_from = v) %>%
  tbl(caption = "Share of each catchment's forecast children expected at its own schools. Source: determination, Table 9.")
```

Longhill's catchment is forecast to retain
`r sprintf("%.0f", min(adj$retention$retained[adj$retention$catchment == "Longhill"]))`–`r sprintf("%.0f", max(adj$retention$retained[adj$retention$catchment == "Longhill"]))`%
of its children at its own school, against
`r sprintf("%.0f", min(adj$retention$retained[adj$retention$catchment == "DS_Varndean"]))`–`r sprintf("%.0f", max(adj$retention$retained[adj$retention$catchment == "DS_Varndean"]))`%
for Stringer and Varndean. Part of that gap is the two city-wide faith schools,
which draw from everywhere; the rest is the outflow above.

::: {.callout-important}
## This is the largest single lever available, and it is not a catchment lever

Roughly `r lh_25` children a year leave the city from one catchment. That is
more than the entire effect of the 2028 boundary change, more than any
admission-number reduction elsewhere in the city, and more than the modelled
effect of the 2024 catchment redraw.

It is also the only quantity in this document whose value does not shrink as
the cohort shrinks, because retaining a child **adds** to the system rather
than moving a child within it. Every other lever tested here redistributes.

What it would take to reverse it is a question about why families in the east
choose to travel out of the city, which no spatial model answers — the journey
times in @sec-network show these are not families being driven out by
distance. It is the question the council's own evidence most clearly raises and
least clearly addresses.
:::

# The 2028 boundary change

On 16 July 2026 the Local Government Secretary confirmed that Brighton & Hove's
boundary will expand eastward from **1 April 2028** to take in East Saltdean,
Telscombe and Peacehaven. Falmer, which had been under consideration, was
dropped.

```{r lgr-map}
#| fig-cap: "The expanded authority. Orange are the four wards joining in 2028. Filled circles are secondary schools inside the new boundary; hollow ones stay outside."

wards_wgs <- st_transform(os$wards, 4326)

pip_sf <- os$pip %>%
  left_join(inp$schools %>% select(name, easting, northing), by = "name") %>%
  filter(!is.na(easting)) %>%
  bind_rows(
    tibble(name = "Seahaven Academy", la = "East Sussex", roll = 769,
           joins = FALSE, dist_km = os$pip$dist_km[os$pip$name == "Seahaven Academy"],
           easting = 543952, northing = 100605)
  ) %>%
  distinct(name, .keep_all = TRUE) %>%
  st_as_sf(coords = c("easting", "northing"), crs = 27700) %>% st_transform(4326)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = catch_sf, fillOpacity = 0.06, color = "#8172B3", weight = 1.5,
              label = ~catchment, group = "Current catchments") %>%
  addPolygons(data = wards_wgs, fillColor = "#DD8452", fillOpacity = 0.35,
              color = "#DD8452", weight = 2, label = ~WD25NM,
              popup = ~paste0("<b>", WD25NM, "</b><br>Joins Brighton & Hove April 2028"),
              group = "Joining in 2028") %>%
  addCircleMarkers(data = pip_sf %>% filter(!joins), radius = 6, color = "grey35",
                   weight = 2, fillOpacity = 0, label = ~name,
                   popup = ~paste0("<b>", name, "</b><br>", la,
                                   "<br><i>Stays outside</i>")) %>%
  addCircleMarkers(data = pip_sf %>% filter(joins), radius = 9, color = "black",
                   weight = 2, fillColor = "#2ca02c", fillOpacity = 0.9,
                   label = ~name,
                   popup = ~paste0("<b>", name, "</b><br>", la,
                                   "<br><b>Inside the new boundary</b>")) %>%
  addLayersControl(overlayGroups = c("Current catchments", "Joining in 2028"),
                   options = layersControlOptions(collapsed = FALSE))
```

```{r lgr-pip}
os$pip %>%
  filter(joins | dist_km < 7) %>%
  transmute(School = name, `Local authority` = la, Postcode = postcode, Roll = roll,
            Verdict = if_else(joins, "JOINS Brighton & Hove",
                              paste0("Stays outside — ", sprintf("%.2f km", dist_km),
                                     " from the new boundary"))) %>%
  tbl(caption = "Secondary schools and the 2028 boundary (point-in-polygon test)") %>%
  row_spec(1, bold = TRUE, background = "#e8f5e9")
```

A point-in-polygon test against the four ward boundaries settles a question
that is often muddled: **Peacehaven Community School is inside. Seahaven
Academy is not** — it sits in Newhaven,
`r sprintf("%.2f km", os$pip$dist_km[os$pip$name == "Seahaven Academy"])` beyond
the eastern edge, and none of the four wards reaches it. The two towns are
adjacent; only one is moving.

So Brighton & Hove acquires **one** secondary school and roughly
`r dem("Expansion area", 2026)` Year 7 children a year. Peacehaven Community
School has `r fmt(os$pip$roll[os$pip$name == PCHS])` pupils against a GIAS
capacity of 890 — it is full. **The city is not gaining an unserved population;
it is gaining a served one, with its school attached.**

# Is there a viable school at Ovingdean?

## The money

```{r lh-finance}
os$lh_fin %>%
  transmute(Year = year_label, Roll = roll,
            `Income per pupil` = paste0("£", fmt(income_pp)),
            `Spend per pupil` = paste0("£", fmt(expenditure_pp)),
            `In-year balance` = paste0("£", fmt(balance)),
            `Revenue reserve` = paste0("£", fmt(reserve))) %>%
  tbl(caption = "Longhill High School, from the published DfE financial returns")
```

Longhill is not short of money per pupil. Its income per pupil rose to
**£`r fmt(os$lh_fin$income_pp[nrow(os$lh_fin)])`**, among the highest of any
secondary school in the city, precisely *because* its roll is falling faster
than its funding. What it cannot do is spend less: it has run an in-year deficit
for three consecutive years and its reserves have fallen from
£`r fmt(os$lh_fin$reserve[1])` to £`r fmt(os$lh_fin$reserve[nrow(os$lh_fin)])`.

**At the average burn of £`r fmt(round(os$reserve_burn))` a year, the remaining
reserve lasts about `r sprintf("%.0f", os$years_left)` more years** — around
`r round(2025 + os$years_left)`, the year after the boundary change takes
effect and four years before its first Year 7 cohort would reach Year 11.

```{r size-finance}
#| fig-cap: "Financial health by school size, all English state secondaries, 2021-22 to 2024-25."
#| fig-height: 4

os$size_fin %>%
  ggplot(aes(size_band, pct_deficit)) +
  geom_col(fill = "#c44e52", width = 0.68) +
  geom_text(aes(label = paste0(round(pct_deficit), "%")), vjust = -0.4, size = 3.3) +
  expand_limits(y = 42) +
  labs(title = "Smaller secondary schools are far more likely to run a deficit",
       subtitle = paste0("Share of school-years with a negative in-year balance, ",
                         fmt(sum(os$size_fin$n)), " school-years"),
       x = "Pupils on roll", y = "% in deficit")
```

This is not unique to Longhill. Across `r fmt(sum(os$size_fin$n))` school-years
of English secondary finance returns, schools under 500 pupils run a deficit
`r round(os$size_fin$pct_deficit[os$size_fin$size_band == "<500"])`% of the time
against about
`r round(median(os$size_fin$pct_deficit[os$size_fin$size_band != "<500"]))`% for
everyone else — despite receiving more per pupil. A secondary school staffs a
maths, science and languages department whether its year group is 90 or 240.

## What size of school does the demand support?

```{r pan-envelope}
#| fig-cap: "How often Longhill recruits to each admission number, across the full parameter sweep. Note this is the share of model RUNS reaching 90% of PAN, not a fill rate - see the caution below."
#| fig-height: 4

env$q1 %>%
  pivot_longer(-entry_year, names_to = "pan", values_to = "pct") %>%
  mutate(pan = factor(parse_number(pan))) %>%
  ggplot(aes(pan, factor(entry_year), fill = pct)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = paste0(round(pct), "%")), size = 3.6) +
  scale_fill_gradient2(low = "#d73027", mid = "#fee08b", high = "#1a9850",
                       midpoint = 50, limits = c(0, 100), guide = "none") +
  labs(title = "How often does Longhill recruit to its number?",
       subtitle = paste0("Share of ", fmt(nrow(env$envelope)),
                         " model runs in which the school reaches 90% of PAN"),
       x = "Published Admission Number", y = "Year 7 entry year")
```

This is the central result, and it does not depend on any assumption the reader
has to accept. Across the **entire** parameter sweep:

- At the PAN of 210 now in force, Longhill reaches 90% of its number in
  **`r round(env$q1[["PAN 210"]][env$q1$entry_year == 2035])`%** of runs by
  2035. At the 240 it carried until 2026, that falls to
  **`r round(env$q1[["PAN 240"]][env$q1$entry_year == 2035])`%**.
- At 150, in `r round(env$q1[["PAN 150"]][env$q1$entry_year == 2035])`%.
- At 120, in `r round(env$q1[["PAN 120"]][env$q1$entry_year == 2035])`%.

**Neither 240 nor the 210 now in force is sustainable under most defensible
parameterisations.**

::: {.callout-caution}
## This result differs from the calibrated model — and the difference changes the recommendation

**Read the chart above carefully: it does not show fill rates.** Each cell is
the share of model *runs* in which Longhill reaches 90% of its number — a
measure of how robust a conclusion is to the unknown parameters, not of how
full the school gets. The equivalent chart in the calibrated analysis shows
actual fill rate, and the two are easy to confuse because they are both
percentages on a red-to-green grid.

Set side by side for 2035, the difference is not subtle:

| PAN | share of runs reaching 90% (above) | modelled fill rate, calibrated |
|---|---|---|
| 120 | 97% | **72%** |
| 150 | 90% | **57%** |
| 180 | 80% | **48%** |
| 240 | 50% | **36%** |

**The two disagree on the policy conclusion, not merely on the numbers.** This
section would naturally be read as saying a PAN of 120 to 150 is sustainable.
On the calibrated model it is not: by 2035 a PAN of 150 fills 57%, and even
120 — already a small school by English standards — reaches only 72%. There is
no admission number at which the calibrated model has Longhill comfortably
full at the end of the window on its current site.

Both the metric and the over-prediction described in @sec-envelope push in the
same direction here, which is why the gap is so wide. **What survives is the
ranking and the exclusion at the top**: 240 and 210 are not viable on either
analysis, and smaller is better. What does not survive is the implication that
120 to 150 is comfortable. On the records it is the least bad option, not a
solution.
:::

# Deprivation and the catchment system {#sec-deprivation}

Everything so far counts children. This section asks which children — the
question *Equity in Education* and *Class Divide* have pressed throughout, and
which none of the modelling above has touched.

IDACI, the Income Deprivation Affecting Children Index, is published by
neighbourhood, so the geography of child poverty in Brighton is fully open.

```{r idaci-map-open}
#| fig-cap: "IDACI by neighbourhood. Darker is more deprived. Circles are secondary schools."

idaci_sf <- dep$zones %>%
  st_as_sf(coords = c("zone_e", "zone_n"), crs = 27700) %>% st_transform(4326)
pal_d <- colorNumeric("magma", domain = idaci_sf$idaci_score, reverse = TRUE)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(data = idaci_sf, radius = 6, stroke = FALSE,
                   fillColor = ~pal_d(idaci_score), fillOpacity = 0.8,
                   label = ~sprintf("IDACI %.3f (decile %d)", idaci_score, idaci_decile)) %>%
  addCircleMarkers(data = st_transform(
                     st_as_sf(inp$schools, coords = c("easting", "northing"),
                              crs = 27700), 4326),
                   radius = 6, color = "black", weight = 2, fillColor = "white",
                   fillOpacity = 0.9, label = ~name) %>%
  addLegend("bottomright", pal = pal_d, values = idaci_sf$idaci_score,
            title = "IDACI", opacity = 0.9)
```

The deprived neighbourhoods sit in the east — Whitehawk, Moulsecoomb, parts of
Woodingdean — and in a band along the seafront. That is the geography every
result in this document has been describing without naming.

## What each catchment starts with

Published data cannot say which school each child attends, so this bundle
cannot compute a school-level intake profile. It can do something arguably more
useful for judging catchment policy: the deprivation of each catchment's own
child population — the distribution the admissions system begins from, and the
thing a redraw directly changes.

```{r dep-open-profile}
dep$profile %>%
  filter(regime == "optionZ") %>%
  transmute(Catchment = catchment, Children = round(children),
            `Mean IDACI` = round(mean_idaci, 3),
            `% in the most deprived 30% of areas` = round(pct_deprived, 1)) %>%
  arrange(desc(`% in the most deprived 30% of areas`)) %>%
  tbl(caption = "Deprivation of each catchment's resident children, under the map in force")
```

BACA's catchment is in a different world from the rest — `r round(dep$profile$pct_deprived[dep$profile$regime == "optionZ" & dep$profile$catchment == "BACA"])`% of
its children live in the most deprived third of neighbourhoods, against
`r round(dep$profile$pct_deprived[dep$profile$regime == "optionZ" & dep$profile$catchment == "Hove_Blatch"])`% in Hove and Blatchington's.

::: {.callout-important}
## The 2024 redraw moved deprivation out of Longhill's catchment

The same calculation under both maps shows what the change did
distributionally.

```{r dep-open-redraw}
dep$comparison %>%
  transmute(Catchment = catchment,
            `Pre-2024 %` = round(pre2024, 1),
            `Option Z %` = round(optionZ, 1),
            Change = round(change, 1)) %>%
  tbl(caption = "Share of each catchment's children in the most deprived 30% of areas")
```

**Longhill's catchment falls from
`r round(dep$comparison$pre2024[dep$comparison$catchment == "Longhill"], 1)`% to
`r round(dep$comparison$optionZ[dep$comparison$catchment == "Longhill"], 1)`% deprived —
a drop of `r abs(round(dep$comparison$change[dep$comparison$catchment == "Longhill"], 1))` points — while Stringer and Varndean's rises by
`r round(dep$comparison$change[dep$comparison$catchment == "DS_Varndean"], 1)`.**
Every other catchment moves by under a point.

That is the redraw doing precisely what it was designed to do. Moving Whitehawk
into the Stringer and Varndean catchment transfers deprived children towards
the city's higher-attaining schools, and the arithmetic confirms it works in
those terms.

It is also, in the same stroke, what @sec-redraw shows costs Longhill: the
children moved out are the ones who chose it. The equity gain and the viability
loss are two descriptions of one transfer.
:::

## How segregated is Brighton, really?

The Gorard Segregation Index summarises how unevenly disadvantaged children are
spread: 0 is perfectly even, 1 complete separation. It reads as the proportion
who would have to move for the distribution to be even.

Across catchment populations it is **`r sprintf("%.3f", dep$gorard[["optionZ"]])`**
under the map in force, against `r sprintf("%.3f", dep$gorard[["pre2024"]])`
before. Both are low.

That corroborates *How to Pull the Right Lever* from a different direction: on
this measure Brighton's schools are already well mixed by national standards.
The admissions criterion aimed at the same problem is examined in @sec-fsm-open.

### Segregation of schools, not just of neighbourhoods

The figure above measures where children **live**. What matters for the policy
argument is where they **end up**, and the model produces that — so the index
can be computed on modelled intakes instead. It inherits the model's parameter
uncertainty, so it is swept rather than quoted as a single number.

```{r seg-sweep}
dep$intake_by_gamma %>%
  transmute(`Catchment effect (γ)` = gamma,
            Lowest = round(lo, 3), Median = round(med, 3),
            Highest = round(hi, 3)) %>%
  tbl(caption = paste0("Gorard index across modelled school intakes, ",
                       nrow(dep$intake_sweep),
                       " runs spanning every attractiveness specification and decay value"))
```

Two things stand out.

**School intakes are less segregated than neighbourhoods.** The median across
the sweep is `r sprintf("%.3f", median(dep$intake_sweep$gorard))`, against
`r sprintf("%.3f", dep$gorard[["optionZ"]])` for the catchment populations
those children are drawn from. Choice moves children across boundaries, and on
balance it mixes rather than sorts. That is not the direction the consultation
debate generally assumed.

**But how much it mixes depends almost entirely on the catchment rule.** With
no catchment effect at all the index falls to
`r sprintf("%.3f", dep$intake_by_gamma$med[dep$intake_by_gamma$gamma == 0])`;
at the strongest priority tested it rises to
`r sprintf("%.3f", dep$intake_by_gamma$med[dep$intake_by_gamma$gamma == 2.4])`,
close to the residential figure.

::: {.callout-important}
## The catchment system is what carries residential segregation into schools

That gradient is the substantive result, and it is available from published
data alone.

Brighton's neighbourhoods are unevenly deprived — that is the map at the top of
this section, and no admissions policy changes it. What determines whether that
unevenness reappears **inside schools** is how tightly children are bound to
the area they live in. A strong catchment rule reproduces the residential
pattern; a weak one lets choice dilute it.

So the debate about catchments is, whether or not it has been framed this way,
a debate about how much of the city's residential segregation to import into
its schools. Every argument for tightening catchment priority — reducing
travel, giving families certainty, protecting local schools — is also an
argument for schools that look more like their neighbourhoods. Every argument
for loosening it runs the other way.

Two cautions. The range is wide because γ is one of the parameters open data
cannot pin down (@sec-wj), so this identifies the mechanism rather than
measuring its current strength. And the whole range sits low by national
standards: even at the strongest catchment effect tested, Brighton is not a
segregated school system.
:::
The consultation proceeded as though the city had a segregation problem to
solve; the evidence is of unevenness at the tails — BACA at one end, Hove and
Patcham at the other — rather than a segregated system.

Note also that the index moves *up* slightly with the redraw, not down. That is
not a failure by the policy's own lights: its aim was to give deprived children
access to higher-attaining schools, not to even out where children live. But it
is worth knowing that the residential distribution the system starts from
became marginally less even, not more.

::: {.callout-note}
## What this cannot show

Where children actually end up. This is the deprivation of each catchment's
**resident** children, not of each school's **intake** — and the gap between
the two is exactly the choice behaviour that published statistics do not
record. A school in a deprived catchment whose families largely go elsewhere,
and a school drawing disadvantaged children from across the city, look
identical here.

That is the same origin–destination gap identified in @sec-wj, and it is the
single most valuable thing the council could release.
:::

## What individual records add here, and why it matters

This is one of the places where the restricted data changes the picture rather
than merely sharpening it, so it is worth setting out what it shows — labelled
as unreproducible from open sources rather than presented as a finding of this
document.

Running the same IDACI attachment through individual preference records gives
each school's **intake** profile, not just its catchment's. Two things follow
that nothing above can see.

**The gap between catchment and intake is large, and it runs in opposite
directions.** Longhill's catchment is around 25% deprived on the figures above;
its actual intake is close to 49%. The affluent families in its catchment leave
— to other city schools, and disproportionately out of the city altogether,
where the leavers are the least deprived group of all. Hove Park shows the
same effect within an affluent catchment: it takes a roughly 30% deprived
intake from a catchment whose children are about 15% deprived, because
Blatchington Mill absorbs the families best placed to choose it.

**The mechanism is a choice effect, not an admissions one.** Fitting the weight
families place on school attractiveness separately by deprivation shows the
least deprived third discriminating between schools about twice as sharply as
the most deprived third, with the difference significant at p < 0.0001. Nobody
is excluding poorer children from the popular schools; more advantaged families
are simply better at securing the place they want. Within any catchment, the
less popular school is left with the more deprived intake.

That is the sorting process the segregation literature describes, and it
happens underneath a Gorard index that looks reassuringly low. The index says
Brighton's schools are well mixed **relative to national variation**. It does
not say the mixing is arrived at neutrally.

**Why this belongs in a public document even though it cannot be reproduced
here.** It bears directly on the policy question. A catchment redraw changes
which school a family is *entitled* to; it does not change how well placed they
are to act on a preference. Any policy justified by improving access for
disadvantaged children is working against a sorting mechanism that this open
model cannot see and the council's own published statistics do not measure —
and the council could measure it tomorrow from records it already holds.

**Running the levers through it produces one further result, and it is
uncomfortable.** Each option for Longhill can be scored not just on how many
children it recruits but on which:

- **Relocating to Elm Grove** raises the deprivation of its intake by around
  12 percentage points. The site sits nearer Whitehawk, so the children it
  would newly reach are disproportionately the city's poorest. That is the
  recruitment gain in @sec-redraw seen from the other side.
- **Retaining the families who currently leave the city** lowers it by about 4
  points, because those leavers are the least deprived group in the catchment —
  22% against 43% of the children who actually attend Longhill.
- **Admission-number reductions elsewhere** move it by under a point. They
  redistribute children between schools without changing which children are in
  play.

So **a viable eastern school and a more even spread of disadvantage are in
tension**. Every arrangement that fills Longhill does so by giving it more of
the children who live nearest to it, and those are the poorest in Brighton. The
movements are small on a Gorard index that starts low, and none turns the city
into a segregated system — but the direction is consistent, and it runs against
what the 2024 redraw was designed to achieve.

None of that can be derived from published statistics. It is the clearest case
in this document for why the origin–destination data matters: without it, a
council can pursue a policy for equity reasons and a policy for viability
reasons without ever discovering that the two work against each other.

# King's, Hove Park, and the other end of the city {#sec-kings-open}

Almost everything in this document concerns Longhill and the east. The published application history contains a second pattern, at the opposite end of the city, which is at least as consequential and has had almost none of the attention.

```{r kings-open-series}
#| fig-cap: "First preferences for King's School and Hove Park School, from the council's published allocation factsheets. Nothing here is modelled."

kp <- readRDS(file.path(PUBLIC_OUT, "factsheet_panel.rds"))$factsheets %>%
  filter(grepl("King's|Hove Park", name)) %>%
  mutate(School = if_else(grepl("King", name), "King's School", "Hove Park School"))

kp %>%
  ggplot(aes(year, pref1, colour = School, group = School)) +
  geom_line(linewidth = 1) + geom_point(size = 2.4) +
  scale_colour_manual(values = c("King's School" = "#7570b3",
                                 "Hove Park School" = "#d95f02")) +
  expand_limits(y = 0) +
  labs(title = "The rise of one school tracks the decline of the other",
       subtitle = "First preferences, published allocation factsheets",
       x = "Year 7 entry year", y = "First preferences", colour = NULL) +
  theme(legend.position = "bottom")
```

```{r kings-open-corr}
#| include: false
kw <- kp %>% select(year, School, pref1, off_total) %>%
  tidyr::pivot_wider(names_from = School, values_from = c(pref1, off_total))
kr_pref <- cor(kw[["pref1_King's School"]], kw[["pref1_Hove Park School"]], use = "complete.obs")
kr_off  <- cor(kw[["off_total_King's School"]], kw[["off_total_Hove Park School"]], use = "complete.obs")
khp26 <- round(kw[["off_total_Hove Park School"]][kw$year == 2026])
```

Between 2014 and 2026, King's first preferences rose from 105 to 218 while Hove Park's fell from 181 to 103. Offers moved the same way — King's from 115 to 180, Hove Park from 260 to `r khp26`. **The two series correlate at `r sprintf("%.2f", kr_pref)` on first preferences and `r sprintf("%.2f", kr_off)` on offers.** One school has roughly doubled while the other has roughly halved, over the same twelve years, in the same part of the same city.

That is published data, and it is the whole of what published data can say. The correlation is striking but it is not, by itself, evidence of a mechanism: two schools in the same area with a fixed pool of children will tend to move oppositely whatever the cause, and Hove Park's own circumstances are an equally available explanation.

::: {.callout-note}
## What the records show, and why it matters for any further expansion

The parallel analysis on individual records adds two facts that published data cannot reach.

**King's is in practice a Hove catchment school.** Around 87% of its intake comes from the Hove Park and Blatchington Mill catchment, and it takes roughly **one in five of all the children living in that catchment**. No other catchment loses more than about 3% of its children to it. It admits on faith criteria and holds no catchment of its own, but its draw is overwhelmingly local.

**It takes the least deprived intake in the city** — about 17% of its children from deprived neighbourhoods, below every other secondary in Brighton & Hove. Hove Park, in the same part of the city, is about 30%.

Set that against @sec-single-open, where Hove Park's own catchment geography is the *least* deprived in the city. The ordering of intakes is the exact inverse of the ordering of the ground the schools stand on. Three institutions draw on the same population; two of them take the less deprived children, and Hove Park takes what is left.

**The question this raises is whether further expansion at King's is compatible with a viable Hove Park in the medium term.** In a shrinking cohort every additional place at King's is drawn from a catchment that is itself getting smaller, and — on the evidence above — disproportionately from the less deprived part of it. That reduces Hove Park's roll and concentrates disadvantage in what remains, at the same time.

It is structurally the same problem as Longhill's, and harder. Longhill's levers are catchment, size and site. Hove Park's competitor has no catchment, admits on criteria the council does not set, and cannot be reached by any instrument used elsewhere in this report.

**That analysis has now been done on the records, and it revises the hypothesis.** Families naming King's first turn overwhelmingly to Blatchington Mill (39%) and Cardinal Newman (38%) as their second choice — the other well-regarded school and the other faith school. Hove Park is the second choice of only **12%** of them, down from 23% three rounds earlier. Meanwhile King's is named second by 23% of Hove Park's *own* first-preference families, up from 16%.

So King's is not taking families who would otherwise have filled Hove Park. It competes with Blatchington Mill and Cardinal Newman. What the data shows instead is that **Hove Park is dropping out of choice sets** — families who want King's no longer hold it in reserve, while families who want Hove Park increasingly hold King's in reserve.

That is a harder problem than competition for a fixed pool. Constraining King's would not return those families, because they are not moving between those two schools; they would go to Blatchington Mill, already the dominant second choice for both groups. And it means Hove Park's more deprived intake is a **residual** — what remains after three other institutions have been chosen from across the same catchment — rather than the result of King's skimming its queue.

The policy implication follows: capping King's would not fix Hove Park. The question that matters is why Hove Park is falling out of choice sets across its own catchment, and that is about the school and how it is regarded rather than about admissions arithmetic. No catchment redesign reaches it.

Two limits. The records cover three rounds (2023-24 to 2025-26) while King's grew mainly between 2014 and 2020, so this shows the pattern *now*, not what happened during the expansion. And second preferences are strategic as well as sincere — some of Hove Park's decline as a second choice may be families correctly judging they will be offered it regardless.
:::

## Hove Park is being chosen against on the wrong measure {#sec-hove-park-va}

The section above leaves an obvious question: if Hove Park is dropping out of choice sets across its own catchment, what are families reacting to? Published data answers this one fully.

```{r hp-open-va}
av <- inp$attract %>%
  transmute(School = name, `Attainment 8` = round(att8, 1),
            `Progress 8` = round(p8, 2),
            `Intake-adjusted (Lever)` = round(va_lever, 2),
            `Preferences per place` = round(prefs_per_place, 2)) %>%
  arrange(desc(`Attainment 8`))
av %>% tbl(caption = "Three measures of the same schools, against how strongly families choose them")
```

```{r hp-open-cor}
#| include: false
ia <- inp$attract
rho_of <- function(v) cor(ia$prefs_per_place, ia[[v]], method = "spearman", use = "complete.obs")
r_att <- rho_of("att8"); r_p8 <- rho_of("p8"); r_va <- rho_of("va_lever")
r_att_p8 <- cor(ia$att8, ia$p8, method = "spearman", use = "complete.obs")
r_att_va <- cor(ia$att8, ia$va_lever, method = "spearman", use = "complete.obs")
```

How strongly families choose a school correlates with **Attainment 8 at `r sprintf("%.2f", r_att)`**, with **Progress 8 at `r sprintf("%.2f", r_p8)`**, and with the **intake-adjusted measure at `r sprintf("%.2f", r_va)`**.

The middle figure is the interesting one, and it is a caution about Progress 8 rather than a finding about families. **Progress 8 correlates with Attainment 8 at `r sprintf("%.2f", r_att_p8)`** across these eleven schools — it is very nearly a restatement of the raw measure, so it cannot separate "families choose on attainment" from "families choose on progress". The intake-adjusted measure used in [*How to Pull the Right Lever*](https://adamdennett.github.io/) correlates with Attainment 8 at only `r sprintf("%.2f", r_att_va)`, so it can. And on that measure, families are **not** tracking what schools add.

For Hove Park the consequence is direct. It sits near the bottom on Attainment 8, but on the intake-adjusted measure it is **positive and above Cardinal Newman and Patcham** — two schools chosen substantially more strongly than it is. Brighton Aldridge, which has the lowest Attainment 8 in the city, adds more on this measure than Blatchington Mill, Cardinal Newman or Hove Park.

::: {.callout-warning}
## The trap this creates, and why no catchment policy reaches it

Attainment 8 is very largely a measure of who a school admits rather than what it does with them. That produces a closed loop, and Hove Park is inside it:

1. It receives the residual intake after three other institutions have been chosen from its catchment (@sec-kings-open).
2. A more deprived intake produces a lower Attainment 8, almost mechanically.
3. Families choose on Attainment 8, so it is chosen less.
4. Which makes its intake more residual still.

Nothing in that loop requires the school to be getting worse, and the intake-adjusted measure says it is not. **The loop is driven by the measure families use, and it will keep turning regardless of what the school does.**

This is why the levers in the rest of this report do not reach Hove Park. Catchment, admission number and site all change *who is in the queue*. None of them changes what parents are looking at when they choose. Publishing an intake-adjusted measure alongside the raw one — which any authority could do — is the only intervention here that acts on the actual mechanism.

**One thing this cannot settle**: whether families choose on Attainment 8 because they do not know a contribution measure exists, or because they know and prefer the peer group anyway. Those imply completely different remedies — the first is answerable with better published information, the second is not. It is a survey question rather than a data question, and a council could answer it in a single admissions round.
:::

Eleven schools is eleven observations, and attractiveness, attainment and intake are entangled. The claim worth defending is the ordering one: families' revealed ordering matches raw attainment far more closely than it matches the intake-adjusted contribution, and Hove Park is among the schools where those two orderings diverge most.

# What the 2024 redraw did {#sec-redraw}

Before asking what a new catchment could do, it is worth pricing the one the
city has already adopted. Every run in the sweep was repeated on both maps —
same site, same decay, same attractiveness, same catchment strength, same
entry year — so the redraw can be isolated.

```{r redraw-pop}
inp$regime_pop %>%
  mutate(children = round(children)) %>%
  pivot_wider(names_from = regime, values_from = children) %>%
  transmute(Catchment = catchment,
            `Pre-2024` = pre2024, `Option Z` = optionZ,
            Change = optionZ - pre2024) %>%
  arrange(Change) %>%
  tbl(caption = "Children entering Year 7 in 2026, by catchment, under each map")
```

The headcount barely moves: Longhill's catchment gains about
`r round(with(inp$regime_pop, children[regime=="optionZ" & catchment=="Longhill"] - children[regime=="pre2024" & catchment=="Longhill"]))`
children. What changes is who they are. The zones moving *in* are no closer to
Longhill than those moving *out* — both average about 34 minutes away — but
they are markedly closer to the alternatives:

```{r redraw-access}
mv <- inp$zones %>%
  filter(catch_pre2024 != catch_optionZ,
         catch_pre2024 == "Longhill" | catch_optionZ == "Longhill") %>%
  mutate(dir = if_else(catch_optionZ == "Longhill",
                       "Moves in (Kemptown)", "Moves out (Whitehawk & Marina)"))

inp$costs_now %>%
  filter(name %in% c(LH, "Dorothy Stringer School", "Varndean School")) %>%
  inner_join(mv %>% select(zone, dir, Oi), by = "zone") %>%
  group_by(dir, name) %>%
  summarise(mins = round(weighted.mean(cij, Oi), 1), .groups = "drop") %>%
  pivot_wider(names_from = dir, values_from = mins) %>%
  rename(School = name) %>%
  tbl(caption = "Population-weighted journey time from the areas the redraw moves")
```

A child moved into Longhill's catchment has Stringer and Varndean roughly
eight and seven minutes closer to hand than a child moved out of it. The model
can see that, and prices the redraw accordingly:

```{r redraw-effect}
env$q6 %>%
  filter(site == "now") %>%
  group_by(entry_year) %>%
  summarise(Median = round(median(shift), 1),
            Lowest = round(min(shift), 1),
            Highest = round(max(shift), 1),
            `% of runs worse` = round(100 * mean(shift < 0)),
            .groups = "drop") %>%
  rename(`Entry year` = entry_year) %>%
  tbl(caption = "Change in Longhill's modelled recruitment under option Z, against the pre-2024 map")
```

On open data the effect is small and its sign depends on how much weight
catchment priority is given: with no catchment effect at all it is zero by
construction, and at the strongest priority tested it is mildly positive,
because the incoming children gain a priority they did not have. The honest
summary is that **the open model cannot resolve this question** — the shifts
are a fraction of a child either way, well inside the noise of everything else
being swept.

That is itself worth reporting. The redraw is exactly the kind of change whose
effect depends on how families actually behave, not on how far apart things
are, and behaviour is the thing published statistics do not record. An
origin–destination matrix of preferences by area would settle it immediately.

# Shrink, move, or redraw?

::: {.callout-caution}
## This result differs from the calibrated model

The fill rates below inherit the over-prediction described under the chart in this section: because the open model recruits roughly twice as many children to Longhill as the records support, **every fill rate here is too high**, and the PAN at which the school appears viable is correspondingly too large.

**With access to the records, the pattern would change in a specific way.** The calibrated model puts configuration A at 37% by 2035 rather than anything approaching viable, and finds only one configuration — Elm Grove at PAN 120 with redrawn catchments — holding above 80% across the whole window. On the open figures several configurations look survivable at PAN 150. **They are not, on the calibrated model.**

What survives the correction is the *comparative* finding, and it is the one the policy question turns on: moving and redrawing are worth little separately and considerably more together, and no configuration rescues a PAN of 210. Read the rank order, discount the levels.
:::

```{r cfg-table}
os$lh_band %>%
  filter(entry_year %in% c(2026, 2030, 2035)) %>%
  mutate(txt = sprintf("%.0f (%.0f–%.0f)", 100*pmin(central,1),
                       100*pmin(lo,1), 100*pmin(hi,1))) %>%
  select(config, entry_year, txt) %>%
  pivot_wider(names_from = entry_year, values_from = txt) %>%
  rename(Configuration = config) %>%
  tbl(caption = "Longhill fill rate (%) — central specification, with the range across the parameter band in brackets")
```

```{r cfg-plot}
#| fig-cap: "Longhill's natural recruitment under each configuration at the central specification. Dotted lines are whole forms of entry."

os$natural %>%
  mutate(lab = str_remove(config, "^[A-G]\\. ")) %>%
  ggplot(aes(entry_year, natural, colour = lab, group = lab)) +
  geom_hline(yintercept = c(90, 120, 150, 180), linetype = "dotted", colour = "grey60") +
  annotate("text", x = 2035.3, y = c(90, 120, 150, 180),
           label = c("3 FE", "4 FE", "5 FE", "6 FE"), hjust = 0, size = 3, colour = "grey45") +
  geom_line(linewidth = 0.9) + geom_point(size = 2.2) +
  scale_x_continuous(breaks = os$years, limits = c(2026, 2038)) +
  scale_colour_brewer(palette = "Set2") +
  expand_limits(y = 0) +
  labs(title = "Longhill's natural size under each option",
       subtitle = "Central specification; recruitment with the school's own admission number unbinding",
       x = "Year 7 entry year", y = "Children recruited", colour = NULL) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 4))
```

::: {.callout-caution}
## This chart is the most optimistic thing in the report, and it is wrong in a known direction

**Read the level on this chart with great care.** It is the single largest divergence between this open model and the calibrated one, and the divergence is not small.

On records, Longhill's natural recruitment in 2035 is around **79 children** where it stands today. This chart puts it at **`r cnat("A", 2035)`** — roughly double. The whole family of curves sits about twice as high as the calibrated model produces, so where this chart shows Longhill supporting five forms of entry, the calibrated model shows fewer than three.

**Why.** The open model can see how reachable Longhill is but has nothing that measures how strongly families *avoid* it. Published data records where children were offered places, not what they wanted; Longhill's revealed attractiveness on records is the lowest in the city, and this model has no way to know that. Once routed journey times let it see that Longhill is genuinely accessible, it sends far more children there than the city actually does.

**What would change with access to the records.** Every curve on this chart would move down, by roughly a factor of two, and the conclusion would harden considerably: the calibrated model finds no configuration in which Longhill recruits above four forms of entry after 2030, and configuration A falls below three. The *ordering* of the configurations is largely preserved — relocation and redrawing still help, in the same rank order — so the comparative reading survives. **The level does not.**

Treat the shape of this chart as informative and the height of it as an upper bound.
:::

::: {.callout-important}
## Relocation helps — a correction to an earlier version

At the central specification the school recruits `r cnat("A", 2035)` children
in 2035 where it is, against `r cnat("C", 2035)` at Elm Grove. Across the full
sweep in @sec-envelope, relocation has a median effect of
**+`r round(median(env$q3$gain[env$q3$entry_year == 2035]), 1)` children by
2035**, positive in
`r round(100 * mean(env$q3$gain[env$q3$entry_year == 2035] > 0))`% of runs.

An earlier version of this document reported the opposite — that no case for
relocation was visible in open data. **That was wrong, and the cause was the
travel matrix.** There were no routed journey times for the Elm Grove site, so
it was given the average accessibility of a Brighton secondary. Rebuilding the
`r5r` network over a merged East and West Sussex extract, and changing nothing
else, moves the relocation effect from about −4 children to about +58 in 2026.
The site sits on a main corridor and is markedly better connected than the
average the approximation assumed.
:::

What other schools reducing their admission numbers would do is examined in @sec-pan-open.

**Shrinking still works on its own.** Configuration B — a PAN of 150 at
Ovingdean, no move, no redraw — fills in 2026 and holds at `r cfill("B", 2035)`
in 2035. Configuration G, a PAN of 120 at Elm Grove with redrawn catchments,
holds at `r cfill("G", 2035)`.

::: {.callout-warning}
## Read the open relocation figures with caution

Routed costs improved the model's realism but made its fit to published offers
*worse*, not better: R² falls to
`r sprintf("%.2f", CalcRSquared(os$val$observed, os$val$modelled))`, and it now
over-predicts Longhill by
`r round(os$val$modelled[os$val$name == LH] - os$val$observed[os$val$name == LH])`
children against
`r round(os$val$observed[os$val$name == LH])` actually offered.

The reason is visible in the arithmetic. Once the model can see how reachable
Longhill really is, it sends far more children there than the city does. What
it lacks is any measure of how strongly families *avoid* the school —
something only individual preference data supplies. The open model therefore
**overstates Longhill's viability**, and the relocation gain above is an upper
bound rather than an estimate.

The direction of the correction is solid; the magnitude is not.
:::

## Designing catchments rather than guessing them {#sec-design-open}

The configurations above take the catchment map as given. A companion exercise, `public/R/08_catchment_design_open.R`, designs it instead, which matters because Peacehaven Community School joins the city's arrangements and its catchment has to be drawn from scratch rather than inherited.

Assigning each neighbourhood to its nearest school gives a Voronoi diagram, which ignores capacity — the popular central schools would be handed far more children than they have places for. Adding the capacity constraint turns this into optimal transport. Give each catchment a scalar *price*, assign every neighbourhood to whichever minimises **journey time + price**, then raise the price of oversubscribed catchments and lower it for undersubscribed ones until each holds the right number of children. The result is an additively weighted Voronoi diagram, or *power diagram*: it minimises total travel time subject to every catchment being the right size, and its regions are contiguous, so the output is a map rather than a scatter of fragments. Every input is published — LSOA child counts, admission numbers from the adjudicator's determination, and the same routed journey times used throughout this document.

Three constraints turned out to be necessary before the output was usable, and they are worth stating because each corrects a failure that looked plausible on the page.

**Targets are capped at the schools' admission numbers.** Scaling targets proportional to PAN and then to the children available can hand a catchment more children than its school can admit. Patcham was the worst case, targeted at 275 against a PAN of 225. A catchment can now come out smaller than its proportional share, which is the honest answer when the school is small.

**Each neighbourhood may only be assigned to one of its three nearest catchments.** Without this the price-adjustment step occasionally satisfies a capacity constraint by reaching across the city.

**The repair step must refuse impossible moves.** After the price adjustment, small detached fragments are absorbed into a neighbouring catchment and boundary neighbourhoods are traded until capacity is restored. Both phases picked the cheapest candidate using functions that *rank* infinities rather than rejecting them — so when the three-nearest rule ruled out every adjacent catchment, a move was still made, to the least impossible option. The symptom was two Peacehaven neighbourhoods sitting in the Dorothy Stringer and Varndean catchment 9.6 km away, with two central neighbourhoods exported to Peacehaven in exchange. The design stage had placed all four correctly; the repair undid it. With the fix, the expansion area is whole within the Peacehaven catchment and that catchment reaches into the city only along the Saltdean fringe.

### Relocation severs Longhill from its eastern hinterland

The designed maps reproduce the finding above — redrawing is worth roughly the same whether or not the school shrinks first, adding of the order of eight points to the 2030 fill rate — but they add one result that the fixed-map configurations cannot show.

```{r cd-open-hinterland}
#| include: false
ocd <- readRDS(file.path(PUBLIC_OUT, "catchment_design_open.rds"))
n_sev <- nrow(ocd$hinterland); c_sev <- round(sum(ocd$hinterland$Oi, na.rm = TRUE))
n_gain <- nrow(ocd$gained);    c_gain <- round(sum(ocd$gained$Oi, na.rm = TRUE))
t_now <- ocd$east_times$`Longhill High School`[ocd$east_times$site == "Ovingdean"]
t_elm <- ocd$east_times$`Longhill High School`[ocd$east_times$site == "Elm Grove"]
```

Hold the admission number at the adjudicator's 210 and design the catchment around each candidate site in turn. **`r n_sev` neighbourhoods, holding `r c_sev` children, belong to Longhill at Ovingdean but not at Elm Grove.** For those places the journey to Longhill goes from an average of `r sprintf("%.0f", t_now)` minutes to `r sprintf("%.0f", t_elm)` — while Dorothy Stringer and Varndean stay where they are, at around 68 and 60 minutes. They are not moving towards a better school; the school is moving away from them, and the designer reassigns them because it has to.

Against that, relocation gains Longhill `r n_gain` neighbourhoods and `r c_gain` children nearer the city centre. So the move is a genuine trade rather than a straight loss — but it is a trade, and it is one the fill rates do not show.

This is a cost of relocation that the fill rates hide, since they are dominated by the central demand the move buys. A viable school serving the east of the city with its natural catchment intact argues for the Ovingdean site and a lower admission number; the highest fill rate argues for Elm Grove and a smaller school still. Those point in opposite directions, and choosing between them is a judgement about what the school is for.

### One catchment per school {#sec-single-open}

Brighton pairs two of its catchments — Hove Park with Blatchington Mill, Dorothy Stringer with Varndean. Splitting them gives each school its own geography. It reassigns every neighbourhood in the city rather than the sixty or so a paired redraw moves, because separating two schools that share a catchment forces a boundary through the middle of an existing one and the price adjustment propagates that outward. So it is a thought experiment, not a proposal — but it answers two questions the paired designs cannot.

**It does not shorten the journey to school.**

```{r cd-open-travel}
ocd$travel_cmp %>%
  transmute(Design = design, Specification = spec, `Longhill site` = site,
            Catchments = catchments,
            `Mean journey (min)` = round(mean_minutes, 2),
            `vs current map` = sprintf("%+.2f", vs_current)) %>%
  tbl(caption = sprintf("Mean journey to the catchment school under each design. The current map gives %.2f minutes.", ocd$baseline_travel))
```

No single-school design beats the best paired one. A paired catchment lets the design send a neighbourhood to whichever of two schools is nearer; splitting the pair removes that freedom, and the journey gets slightly longer. But the honest headline is the size of the spread: **every design in the table sits within about a minute of every other, on a mean journey of `r round(ocd$baseline_travel)` minutes.** Catchment geometry is not where travel-time gains are. Where a school *is* matters; which catchment a child is put in barely does.

**What it would do to deprivation is more interesting**, because this is what the policy debate turns on.

```{r cd-open-dep}
ocd$spec_gorard %>%
  transmute(Specification = spec, Catchments = catchments,
            `Gorard index` = round(gorard, 3),
            `Most − least deprived (pp)` = round(range_pp, 1)) %>%
  tbl(caption = "Segregation of where children live, by catchment specification. Higher Gorard means disadvantage is spread more unevenly across catchments.")
```

Two measures move in opposite directions, and both readings are correct.

The **gap between the most and least deprived catchment narrows sharply** — from `r round(ocd$spec_gorard$range_pp[ocd$spec_gorard$spec == "Current map"])` percentage points under the current map to `r round(ocd$spec_gorard$range_pp[ocd$spec_gorard$spec == "One per school, Ovingdean"])` with one catchment per school. Smaller catchments cannot span the whole social gradient, so the extremes come in.

But the **Gorard index rises**, from `r sprintf("%.3f", ocd$spec_gorard$gorard[ocd$spec_gorard$spec == "Current map"])` to `r sprintf("%.3f", ocd$spec_gorard$gorard[ocd$spec_gorard$spec == "One per school, Ovingdean"])`. More catchments means more of them can be socially distinctive, and the index measures unevenness across all of them rather than the range between the extremes. If the worry is that some children live in a catchment unlike the rest of the city, splitting helps. If the worry is the overall evenness of the spread, it does not.

```{r cd-open-pairs}
ocd$pairs_split %>%
  transmute(School, Children = children, `% deprived` = pct_deprived) %>%
  tbl(caption = "What splitting the two paired catchments would give each school — children living in the catchment, not modelled intakes")
```

Each pair splits into a less deprived and a more deprived half, Hove Park taking the less deprived side of its pair and Varndean the less deprived side of its own. That is worth setting against what @sec-deprivation shows about intakes: the sorting that happens *inside* a shared catchment does not follow the geography underneath it, and pairing masks a social boundary rather than creating one.

### The fairness argument for splitting them {#sec-dual-fairness-open}

Public argument during the consultation held that having some dual and some single-school catchments is unfair, because families in a dual catchment get a *choice* of two schools while families in a single-school catchment get none. Two things are worth saying about it.

**In an oversubscribed catchment there is no choice, only chance.** Naming a school does not secure it; where places are short the tiebreaker decides, and the tiebreaker is random. A dual catchment does not confer two choices. It confers one preference and a probability. How often that bites can only be measured on individual preference and offer records, which are not published — work on those records finds it bites hard at one school in the pair and essentially not at all at the others.

**The pairing does redistributive work that splitting would undo.** This is the part the argument misses. A shared catchment mixes two populations the geography would otherwise separate, and the tables above show the direction: each pair splits into a more and a less deprived half, so the pair is holding together a social range that single-school catchments would divide.

That inverts the fairness case. Removing dual catchments would not remove the element of chance — that comes from oversubscription, not from pairing. It would remove the mixing and leave the chance where it was.

::: {.callout-important}
## The west of the city has been the quiet half of this debate

The consultation has been dominated by deprivation in the east, and the numbers justify the attention: BACA's catchment is far more deprived than anywhere else in the city. But catchment-level statistics are averages, and an average over a large paired catchment conceals whatever variation sits inside it. Hove and Portslade's more deprived neighbourhoods are pooled with Hove Park's much less deprived ones and disappear from the published figures.

```{r cd-open-west}
#| include: false
ocur <- ocd$spec_profiles %>% filter(spec == "Current map")
osng <- ocd$spec_profiles %>% filter(spec == "One per school, Ovingdean")
o_hb <- ocur$pct_deprived[ocur$School == "Hove_Blatch"]
o_hp <- osng$pct_deprived[osng$School == "Hove Park School"]
o_bm <- osng$pct_deprived[osng$School == "Blatchington Mill School"]
o_bmn <- round(osng$children[osng$School == "Blatchington Mill School"] * o_bm / 100)
```

Hove Park and Blatchington Mill together are `r round(o_hb, 1)`% deprived — **the least deprived catchment in the city**. Split them and the two halves are `r round(o_hp, 1)`% and `r round(o_bm, 1)`%. The combined figure is not a description of a place; it is an average across a social boundary, and it is the number that reaches the published statistics.

Give Blatchington Mill its own geography and it holds roughly `r o_bmn` children from the most deprived 30% of neighbourhoods nationally. None of them are visible in any catchment-level figure, because they are pooled with Hove Park's. That is a reason to look at the west of the city with the care the east has received, rather than reading a low catchment average as evidence there is nothing there.
:::

::: {.callout-note}
## How far to trust the size of these gaps

An earlier version of this document apportioned each ward's children between its LSOAs by **postcode count**, which assumes every postcode holds the same number of eleven-year-olds. It does not — family housing, student housing and retirement housing have very different child densities, and the difference runs along exactly the lines catchment boundaries follow. That assumption flattened this section almost to nothing: the Hove Park and Blatchington Mill gap came out at 3.5 percentage points.

These figures use ONS small-area population by single year of age instead, which is published and gives each LSOA's real age-11 share. The gap is now `r round(diff(range(ocd$pairs_split$pct_deprived[ocd$pairs_split$School %in% c("Hove Park School", "Blatchington Mill School")])), 1)` points, against about 25 on pupil-level records — still an understatement, but the same phenomenon rather than a different one.

**Two figures in the table above should still be treated as artefacts.** Varndean at `r round(ocd$pairs_split$pct_deprived[ocd$pairs_split$School == "Varndean School"], 1)`% and Longhill's single-school catchment at around 17% are both far from what records show (roughly 25% and 58%). The single-school designs place boundaries the open model has no way to check, and small catchments are sensitive to exactly where those boundaries land. The paired figures, and the Hove Park / Blatchington Mill contrast in particular, are the ones that hold up.
:::

# The Whitehawk question {#sec-whitehawk-open}

::: {.callout-note}
## This section cannot be done on published data

The 2024 redraw moved the Whitehawk area out of Longhill's catchment and into Dorothy Stringer and Varndean's, and moved part of Kemptown in. Whether that helped or harmed Longhill, and whom it helped, is one of the most contested questions in the consultation — and it turns entirely on how families in those specific neighbourhoods behaved.

Published data can show the redraw's *geography* (@sec-redraw) and its effect on catchment populations. It cannot show what it did to choices, because that needs preferences from particular origins.

The parallel analysis finds something that cuts against both sides of the argument. **Kemptown, the area moved *into* Longhill's catchment, was already choosing Longhill at a higher first-preference rate than Whitehawk was while Whitehawk sat inside it.** Being placed in a catchment does not, by itself, make families choose a school — which is consistent with the finding on catchment priority above, and unhelpful to anyone arguing the redraw alone will fix or has broken Longhill's recruitment.

This is precisely the kind of question where the absence of evidence lets the loudest argument win. It is answerable in an afternoon with the records.
:::

# The free school meals priority {#sec-fsm-open}

Brighton & Hove has introduced a free school meals priority into its secondary
oversubscription criteria, and placed it **above catchment residence**:

| | Criterion |
|---|---|
| 3 | Sibling link |
| 4 | Children in catchment eligible for FSM, up to the city average |
| 5 | Other children eligible for FSM, up to the city average |
| 6 | Open admissions (5% quota) |
| 7 | Children in catchment |
| 8 | Other children |

Criteria 4 and 5 lift an eligible child above every catchment resident.
Criterion 5 is the redistributive one: it places a child at a school they do
not live near, ahead of children who do.

Everything in this section is published. The adjudicator required the council
to set out both its 2026 outturn by criterion and its own 2027 projection, so
the effect can be read off the council's arithmetic rather than modelled.

```{r fsm-open-city}
fo <- readRDS(file.path(PUBLIC_OUT, "fsm_criterion_open.rds"))

fo$city %>%
  transmute(Criterion = criterion, Offers = offers,
            `Share of offers` = sprintf("%.1f%%", share)) %>%
  tbl(caption = "City-wide offers by oversubscription criterion, 2026")
```

```{r fsm-open-proj}
fo$projection %>%
  transmute(School = school, Criterion = criterion,
            `2026 outturn` = y2026, `2027 projected` = y2027,
            Change = change) %>%
  tbl(caption = paste0("The three schools the council expects to be oversubscribed. Source: ",
                       fo$source, "."))
```

Across those three schools the FSM criteria account for `r fo$fsm_2026` places
in 2026, falling to `r fo$fsm_2027` projected for 2027 — a drop of
`r sprintf("%.0f%%", 100 * (1 - fo$fsm_2027 / fo$fsm_2026))`.

::: {.callout-important}
## The priority is being cut by nearly half

Eligibility is being narrowed to **Targeted FSM**. The council states that
approximately **`r sprintf("%.0f%%", 100 * fo$targeted_share)` of currently
FSM-eligible pupils will qualify**, and that it "is anticipating a decrease of
`r sprintf("%.0f%%", 100 * fo$stated_decrease)` in pupils eligible for the FSM
admissions priority".

This matters beyond the criterion itself. The council's equalities case for the
wider admissions changes rests on the proposition that displacement is
"unlikely to include children eligible for free school meals as these pupils
have a higher priority which sits above children living in the school's
catchment area" — a protection being substantially withdrawn in the same year,
with no further modelling completed.

Where those children fall to is set by the order above. In-catchment losers
drop to criterion 7. Out-of-catchment losers drop to criterion 8, below every
catchment child, or into competition for the 5% open quota.
:::

## An arithmetic check worth making

The council's stated basis is that 56% of currently eligible pupils qualify.
Two of its three criterion-4 projections follow that. One does not.

```{r fsm-open-check}
fo$check %>%
  transmute(School = school, `2026` = y2026, `2027 projected` = y2027,
            `Implied share` = sprintf("%.0f%%", 100 * implied_share),
            `Expected at 56%` = expected_at_56,
            `Follows stated basis` = if_else(consistent, "yes", "no")) %>%
  tbl(caption = "Criterion 4 projections against the council's own 56% basis")
```

Blatchington Mill falls from 45 to 25, exactly 56%. Dorothy Stringer falls from
35 to 22, near enough. **Varndean rises, from 61 to 64** — where the council's
own basis would give about `r fo$check$expected_at_56[fo$check$school == "Varndean School"]`.

Criteria 4, 5 and 6 all outrank criterion 7, so a discrepancy of this size
directly affects how many places remain for in-catchment children at what the
council's own figures make the most oversubscribed school in the city. It may
have an explanation; it is not one the published material gives, and it is
worth asking about.

::: {.callout-note}
## What this cannot say

Whether the contraction helps or harms any particular school depends on where
the children losing the priority live, which published data does not record.
The direction is arguable either way: the criterion carries eligible children
*away* from their catchment towards the popular central schools, so withdrawing
it should return some of them — but to which schools, and in what numbers, is
exactly the origin–destination question @sec-wj identifies as the central gap.

What is not in doubt is the timing. The protection shrinks by nearly half in
the same admissions round as the sibling and open-admission changes, and the
council confirms it has not modelled the combined effect.
:::

# What if other schools shrink too? {#sec-pan-open}

Every scenario so far treats Longhill as the only school whose size is in
question. A falling cohort presses on all of them at once, and several have
their own reasons to want to be smaller — Cardinal Newman most of all, at 360
by some distance the largest in the city.

This is testable with published data alone: admission numbers are public, and
the model already redistributes displaced demand. The scenarios below use the
adjudicator's binding 2026/27 numbers, which matters, because the council's
proposed reductions at Blatchington Mill and Dorothy Stringer were overturned —
60 central places the council meant to remove stay in the system.

```{r pan-open}
po <- readRDS(file.path(PUBLIC_OUT, "pan_scenarios_open.rds"))

po$longhill %>%
  mutate(natural = round(natural)) %>%
  pivot_wider(names_from = entry_year, values_from = natural) %>%
  rename(Scenario = scenario) %>%
  tbl(caption = "Longhill's natural recruitment if other schools reduce their admission numbers")
```

::: {.callout-warning}
## The open model cannot answer this one, and it is worth saying why

Reducing Cardinal Newman changes Longhill's recruitment by **nothing at all**
in this model — not a small amount, exactly zero. That is not a finding about
Brighton. It is a failure of the open model, and the diagnosis is visible in
the capacity table.

**The calibrated model finds the opposite, and it is one of the larger
divergences in this document.** On records, cutting Cardinal Newman to 270
sends Longhill **+29 children** in 2026; Newman 270 together with Dorothy
Stringer 270 sends **+55**; adding BACA at 150 sends **+72**. Against Longhill's
calibrated 2026 baseline of about 114, that last figure is a **63% increase** —
the single most effective lever tested anywhere in either report, and larger
than relocation and redrawing combined.

So this section understates a real mechanism to zero. If the cascade matters
as much as the records suggest, the size of *other* schools is a more powerful
instrument for Longhill than anything done to Longhill itself — which, if true,
reframes the policy question substantially. **Published data cannot establish
it, and this report should not be read as evidence against it.**

```{r pan-open-slack}
po$slack %>%
  transmute(School = name, PAN = pan, Intake = round(intake),
            `Spare places` = round(spare)) %>%
  arrange(desc(`Spare places`)) %>%
  head(4) %>%
  tbl(caption = "Where the open model leaves capacity unused, 2026")
```

The open model has Cardinal Newman recruiting only
`r round(po$slack$intake[po$slack$name == "Cardinal Newman Catholic School"])`
of its `r po$slack$pan[po$slack$name == "Cardinal Newman Catholic School"]`
places. Reducing its admission number to 270 therefore binds on nothing,
because the school never reaches 270 in the first place.

In reality Cardinal Newman fills every year and is the most oversubscribed
school in the city. The open model under-predicts it because the only public
signal of a faith school's popularity is its first-preference count, and that
badly understates demand for a school which families know admits on religious
criteria — many who would want a place never express the preference. The same
weakness that makes @sec-choice cautious about attractiveness proxies bites
hardest here.

**So this question needs the restricted data.** Individual preference records
show Cardinal Newman as by far the most attractive school in Brighton, full at
360, and a reduction to 270 displacing children who cascade towards the two
schools with room. This bundle cannot reproduce that, and rather than present
a table of zeroes as though it were a result, the honest statement is that the
open model is blind to it.

It is the clearest single example in this document of what an
origin–destination matrix of preferences would add.
:::

# How much does any of this depend on unknowns? {#sec-envelope}

```{r envelope-plot}
#| fig-cap: "Longhill's natural intake across every combination swept. Each line is one attractiveness specification at one catchment-effect strength."

env$envelope %>%
  filter(site == "now") %>%
  ggplot(aes(beta, natural_intake, colour = w_label,
             group = interaction(w_label, gamma))) +
  geom_hline(yintercept = env$pans, linetype = "dotted", colour = "grey65") +
  geom_line(alpha = 0.55, linewidth = 0.6) +
  facet_wrap(~ entry_year, nrow = 1) +
  scale_colour_brewer(palette = "Dark2") +
  expand_limits(y = 0) +
  labs(title = "The full parameter envelope",
       subtitle = "Dotted lines are candidate admission numbers of 120, 150, 180, 210 and 240",
       x = "Distance decay (beta)", y = "Children recruited", colour = NULL) +
  theme(legend.position = "bottom")
```

```{r importance}
env$importance %>%
  filter(term != "Residuals") %>%
  mutate(term = recode(term,
    "w_spec" = "Which attractiveness specification",
    "factor(entry_year)" = "Which year",
    "factor(gamma)" = "Strength of catchment priority",
    "factor(beta)" = "Distance decay (beta)",
    "site_label" = "Site (Ovingdean or Elm Grove)")) %>%
  transmute(Input = term, `Share of variation (%)` = round(pct, 1)) %>%
  tbl(caption = "What the answer is actually sensitive to")
```

::: {.callout-note}
## What restricted data would and would not add

The single most consequential unknown is **not** how sharply families discount
distance. Distance decay accounts for
`r round(env$importance$pct[env$importance$term == "factor(beta)"], 1)`% of the
variation in the answer. **How attractive each school is, net of where it sits,
accounts for `r round(env$importance$pct[env$importance$term == "w_spec"], 1)`%.**

That is a specific and modest ask of the council. What would most improve this
analysis is an **origin-destination matrix of Year 7 preferences and offers, by
area of residence and destination school**. Aggregated to catchment or ward
level it carries no disclosure risk, and it would replace the widest band of
uncertainty here with a measurement.
:::

# What this shows, and what it does not

**What holds without any modelling assumption at all**

- Longhill has the lowest potential accessibility to the city's children of any
  secondary school in Brighton (@sec-brightopia), at
  `r round(bright$at_original_beta$access_index[bright$at_original_beta$name == LH])`
  against a city average of 100 — and it has the fewest competing schools
  nearby, so that is a reach problem rather than a crowding one.
- The Elm Grove site is better placed at every distance decay tested. This is
  the one conclusion here that requires no view on attractiveness, admissions
  priority, or family behaviour.

**What holds across the whole envelope**

- Longhill cannot sustain the PAN of 210 now in force into the 2030s: it
  reaches 90% of that number in only
  `r round(env$q1[["PAN 210"]][env$q1$entry_year == 2035])`% of runs by 2035,
  and that is on a model already biased in its favour. At the 240 it carried
  until 2026, `r round(env$q1[["PAN 240"]][env$q1$entry_year == 2035])`%.
- A PAN of 120–150 is defensible.
- The 2028 boundary change does not alter this. It adds one school and its own
  children, and those children live closer to that school than to Longhill.
- The 2024 catchment redraw does not alter it either, in the direction the
  open model can see (@sec-redraw). Whether it makes Longhill's position
  slightly better or slightly worse is beyond what published data can settle.
- Demand tracks published Attainment 8, not value-added.
- Relocating to Elm Grove increases recruitment, in
  `r round(100 * mean(env$q3$gain[env$q3$entry_year == 2035] > 0))`% of runs by
  2035. The direction is robust; the size is not, because the model
  over-predicts Longhill to begin with.

**What this analysis cannot settle**

- *How much* relocation would help. The open model over-predicts Longhill
  badly, so its estimate of the gain is an upper bound.
- What happens to children academically. This models where they would go, not
  what happens when they get there — for which see
  [school_attainment_tool](https://github.com/adamdennett/school_attainment_tool),
  which finds absence a far more powerful lever than any redistribution of
  intake.
- Anything below catchment level about who goes where.

**Standing limits**

- Travel times are now routed throughout, including the expansion area and the
  Elm Grove site. The `r5r` network was rebuilt over a merged East and West
  Sussex extract; no journey time in this document is inferred from
  straight-line distance.
- The model over-predicts Longhill by
  `r round(os$val$modelled[os$val$name == LH] - os$val$observed[os$val$name == LH])`
  children against published offers, so its figures are generous to the school
  and its relocation gain is an upper bound.
- Ten destinations means anything about *why* a school is attractive rests on
  ten observations.
- Only the schools inside the authority are modelled as destinations. Routed
  times to Seahaven, Priory Lewes and Seaford Head now exist but are not yet
  used, so eastern children have nowhere to go in the model except Brighton or
  Peacehaven — which overstates the inflow to the city.

# What the council's own data would settle {#sec-council-data}

This document is built entirely from published sources, and that constraint has been productive: it forces every claim to be checkable by anyone. But it has been binding at specific, identifiable points, and it is worth naming them — partly so readers know where the analysis is weakest, and partly because the data that would fix them already exists.

The council holds individual preference and offer records: for each applicant, the schools named in order, the criterion under which a place was allocated, and the home postcode. Nothing below needs anything the admissions process does not already generate.

**Where the open model is demonstrably wrong, and would stop being.** The validation in @sec-envelope shows this model over-predicting Longhill by a wide margin once routed journey times are used. The reason is structural: the model can see how reachable a school is, but nothing in published data records how strongly families *avoid* one. Preference records supply exactly that, as a school-level term fitted rather than assumed. Until then every relocation figure here is an upper bound, and the report says so repeatedly because it is the single largest source of error in it.

**Whether catchment priority is worth anything.** This document has to assume a value for how much being in catchment shifts a family's choice, and sweeps it across a wide band because the assumption cannot be tested openly. Records settle it directly, by school: compare offer rates for in-catchment and out-of-catchment applicants who named the same school. Work on the records finds the answer at Longhill is **nothing at all** — essentially everyone who names it is offered a place, in catchment or out — which if true city-wide would mean redrawing Longhill's catchment cannot work through priority, only through what families are willing to name. That is a conclusion with direct policy consequences and it is invisible in published data.

**Choice or chance.** The consultation argument that dual catchments confer "choice" can only be tested against individual preferences. The relevant number is the share of in-catchment applicants who named a school first and were refused it. Records show that at one school in a pair this runs near 30%, and at the others essentially zero — so the "choice" is real at some schools and a lottery at others. Published aggregates cannot distinguish those two situations at all.

**How much deprivation the catchment averages conceal.** @sec-single-open shows the pairing masking a substantial social boundary in Hove, but understating it: about 18 percentage points here against 25 on records. Postcode-level home addresses would remove the approximation entirely, and with it the caveat that currently sits on the most policy-relevant finding in this document.

**Who leaves, and why.** Roughly a fifth of the city's children do not enter its state secondary schools. Published data cannot say whether they go independent, to a faith school, or out of the authority, nor where they live. Records can, and the east-of-the-city pattern that emerges is central to any judgement about Longhill's future.

Five sections of this report exist only to mark where published data runs out: @sec-calibration-open, @sec-rules-open, @sec-heterogeneity-open, @sec-required-open and @sec-whitehawk-open. Each says what cannot be established here and what the records would establish. They are the shape of the offer.

None of this requires new collection, a survey, or a change of practice. It requires access to records the admissions process produces as a by-product, under the arrangements that already govern research use of pupil data.

## Why this is worth doing now rather than later

The case is not that the analysis would be interesting. It is that the city is about to make a series of decisions that will bind for fifteen to twenty years, and is making them one at a time.

Consider what is in front of the council across the next few rounds: Longhill's admission number and possibly its site; Dorothy Stringer's admission number, already once refused by the adjudicator; the free school meals criterion, whose reach is about to be cut by nearly half by a change in national eligibility rules; the catchment map, redrawn in 2024 and not yet settled; and the arrival of four wards and a further secondary school in 2028. Each of these is currently a separate decision with a separate evidence base and a separate consultation.

**They are not separate.** Every result in this document says so. Reducing one school's admission number pushes demand onto its neighbours and the effect concentrates, unevenly, on the school least able to absorb it. Redrawing a catchment changes who is in it, and therefore the social composition of two schools, not one. Relocating a school severs it from part of its hinterland while gaining another part. A model of the whole system is the only way to see any of that before it happens rather than afterwards.

The demographic reality makes this urgent rather than optional. The cohort is falling, the fall is not evenly distributed across the city, and surplus capacity has to land somewhere. A city that decides where it lands, deliberately and once, will get a better answer than one that discovers where it landed after five years of separate decisions each of which seemed reasonable in isolation.

That is what this work is for. It is not an argument for any particular outcome for Longhill, or for any school — the results here point in different directions depending on what the city decides it values, and the report says so wherever that happens. It is an argument for having the conversation with the whole picture in view, and for doing so while there is still time for the answer to matter.

The offer stands: the model, the code, and the analysis are open, and the parts that need records will be done properly and shared if the records are made available.

# Testable predictions

These follow from published data. The council holds the records that would
confirm or refute them.

::: {.callout-caution}
## Prediction 1 is already known to be too high

The calibrated model puts Longhill's 2035 natural recruitment at around **79
children** under current catchments — below the bottom of the range quoted in
prediction 1, not inside it. That is the over-prediction described in
@sec-envelope arriving at a specific number.

The prediction is left as published data states it, because that is what this
document is for and a reader should be able to see exactly where the open
sources lead. But it is the clearest single case in the report of a figure the
records would move, and the direction is downward.
:::

1. Longhill's natural recruitment in 2035 sits between
   `r round(env$q2$min[env$q2$entry_year == 2035 & env$q2$site_label == "Ovingdean"])`
   and
   `r round(env$q2$max[env$q2$entry_year == 2035 & env$q2$site_label == "Ovingdean"])`
   children under current catchments, centred near a median of
   `r round(env$q2$median[env$q2$entry_year == 2035 & env$q2$site_label == "Ovingdean"])` children.
2. Relocating to Elm Grove without redrawing catchments *increases* Longhill's
   intake, by a median of
   `r round(median(env$q3$gain[env$q3$entry_year == 2035]))` children a year by
   2035 on this model — which, given the model's known bias towards the school,
   should be read as an upper bound.
3. The four wards joining in 2028 contribute fewer than 10 additional children a
   year to Longhill, under arrangements retaining a catchment priority at
   Peacehaven Community School.

If any is wrong, the data to show it exists. We would welcome the correction.

# Sources {#sec-sources}

| Input | Source |
| --- | --- |
| Travel times | `r5r` routing over OpenStreetMap and Brighton & Hove GTFS, May 2024 |
| Admissions, PAN, preferences | Brighton & Hove City Council published allocation factsheets, 2013–2026 |
| School characteristics | DfE [Get Information About Schools](https://get-information-schools.service.gov.uk/) |
| Attainment, Ofsted, finance | DfE performance tables, via [school_attainment_tool](https://github.com/adamdennett/school_attainment_tool) |
| Population | ONS small area population estimates, mid-2022 |
| Boundaries | ONS Open Geography Portal; council catchment map |
| Reorganisation | [MHCLG decision letters](https://www.gov.uk/government/publications/local-government-reorganisation-decision-letters-to-council-leaders), 16 July 2026 |

```r
source("public/run_public.R")
quarto::quarto_render("public/bh_school_sim_open.qmd")
```
